---
title: "Advice Outcome Analysis"
author: "Group 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

source("scripts/00-packages.R")
source(here("scripts", "01-get-data.R"))
```

## Purpose

This analysis addresses RQ3:

> **Do schools with similar test results give similar secondary-school advice?**

The final visualization focuses on whether school disadvantage (`schoolweging`) changes the relationship between school-level test performance and the percentage of pupils receiving **HAVO-or-higher advice**.

For this draft:

- privacy-suppressed counts (`"<5"`) are replaced by **2.5**;
- split-track advice (`HAVO_VWO`) contributes **0.5** to the HAVO-or-higher outcome;
- test performance is the average percentage reaching **maths 1S** and **reading 2F**;
- `schoolweging` remains continuous in the statistical model;
- the 25th, 50th, and 75th percentiles of `schoolweging` are used only to visualize representative levels of disadvantage.

## 1. Prepare the analysis data

```{r prepare-data}
names(schooladviezen) <- gsub("\\u00a0", "", names(schooladviezen))

advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)

to_num <- function(x) {
  if_else(x == "<5", 2.5, suppressWarnings(as.numeric(x)))
}

adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

advice_data <- schooladviezen |>
  mutate(
    inst = adv_ids[[1]],
    vest = adv_ids[[2]],
    across(all_of(advice_cols), to_num),
    n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice
  ) |>
  filter(n_advice > 0) |>
  select(inst, vest, pct_havo_plus)

reference_data <- referentieniveaus |>
  mutate(
    inst = ref_ids[[1]],
    vest = ref_ids[[2]],
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  ) |>
  select(inst, vest, pct_test)

weighting_data <- schoolweging |>
  transmute(
    inst = sub("\\|.*$", "", OVT),
    schoolweging
  ) |>
  filter(!is.na(schoolweging)) |>
  distinct(inst, .keep_all = TRUE)

model_data <- advice_data |>
  inner_join(reference_data, by = c("inst", "vest")) |>
  inner_join(weighting_data, by = "inst") |>
  filter(!is.na(pct_test), !is.na(pct_havo_plus))

nrow(model_data)
```

The analysis is performed at the **school level**. The outcome is the percentage receiving HAVO-or-higher advice, while test performance is based on standardized reference-level attainment rather than provider-specific raw scores.

## 2. Model the relationship

The interaction model tests whether the relationship between test performance and HAVO-or-higher advice changes with school disadvantage.

```{r moderation-model}
model_havo <- lm(
  pct_havo_plus ~ pct_test * schoolweging,
  data = model_data
)

model_results <- tibble(
  r_squared = summary(model_havo)$r.squared,
  interaction = coef(model_havo)[["pct_test:schoolweging"]]
)

model_results
```

A negative interaction means that the positive relationship between test performance and HAVO-or-higher advice becomes weaker as school disadvantage increases.

### Representative predictions

The statistical model keeps `schoolweging` continuous. The three lines in the visualization represent its 25th, 50th, and 75th percentiles.

```{r predictions}
sw_values <- quantile(
  model_data$schoolweging,
  probs = c(0.25, 0.50, 0.75),
  na.rm = TRUE
)

prediction_data <- expand_grid(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  ),
  schoolweging = as.numeric(sw_values)
) |>
  mutate(
    disadvantage = factor(
      schoolweging,
      levels = as.numeric(sw_values),
      labels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    ),
    fit = predict(model_havo, newdata = pick(pct_test, schoolweging))
  )
```

To make the widening gap easy to interpret, compare the lower- and higher-disadvantage predictions at a few test-performance values.

```{r gap-check}
gap_check <- expand_grid(
  pct_test = c(25, 40, 55, 70),
  schoolweging = c(sw_values[[1]], sw_values[[3]])
) |>
  mutate(predicted = predict(model_havo, newdata = pick(everything()))) |>
  arrange(pct_test, schoolweging) |>
  group_by(pct_test) |>
  summarise(
    lower_disadvantage = first(predicted),
    higher_disadvantage = last(predicted),
    gap = higher_disadvantage - lower_disadvantage,
    .groups = "drop"
  )

gap_check
```

The main pattern for the visualization is the increasing separation between the lower- and higher-disadvantage predictions as test performance rises.

## 3. Interactive visualization

The interactive plot has two states:

1. **Overall:** the observed schools and the overall test-performance relationship.
2. **Show school disadvantage:** schools are coloured by broad disadvantage group and the overall line is replaced by predictions at the 25th, 50th, and 75th percentiles of `schoolweging`.

The broad groups used to colour the points are only a visual aid; the moderation model itself continues to use continuous `schoolweging`.

```{r interactive-data}
model_data <- model_data |>
  mutate(
    disadvantage_group = case_when(
      schoolweging <= sw_values[[1]] ~ "Lower disadvantage",
      schoolweging >= sw_values[[3]] ~ "Higher disadvantage",
      TRUE ~ "Typical"
    ),
    disadvantage_group = factor(
      disadvantage_group,
      levels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    )
  )

high_reference <- prediction_data |>
  filter(disadvantage == "Higher disadvantage") |>
  select(pct_test, high_fit = fit)

prediction_interactive <- prediction_data |>
  left_join(high_reference, by = "pct_test") |>
  mutate(
    difference_high = fit - high_fit,
    hover_text = paste0(
      "<b>", disadvantage, "</b>",
      "<br>Test performance: ", round(pct_test, 1), "%",
      "<br>Predicted HAVO+ advice: ", round(fit, 1), "%",
      "<br>Difference from higher-disadvantage school: ",
      if_else(
        abs(difference_high) < 1e-8,
        "reference",
        paste0(
          if_else(difference_high > 0, "+", ""),
          round(difference_high, 1),
          " percentage points"
        )
      )
    )
  )

overall_model <- lm(pct_havo_plus ~ pct_test, data = model_data)

overall_data <- tibble(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  )
) |>
  mutate(fit = predict(overall_model, newdata = pick(everything())))
```

```{r interactive-plot}
group_colours <- c(
  "Lower disadvantage" = "#2C7BB6",
  "Typical" = "#7F7F7F",
  "Higher disadvantage" = "#D7191C"
)

interactive_plot <- plot_ly()

# Default state: neutral school dots.
interactive_plot <- interactive_plot |>
  add_markers(
    data = model_data,
    x = ~pct_test,
    y = ~pct_havo_plus,
    marker = list(color = "grey55", size = 5, opacity = 0.35),
    hoverinfo = "skip",
    showlegend = FALSE
  )

# Disadvantage-coloured dots, hidden initially.
for (group in levels(model_data$disadvantage_group)) {
  group_data <- model_data |>
    filter(disadvantage_group == group)

  interactive_plot <- interactive_plot |>
    add_markers(
      data = group_data,
      x = ~pct_test,
      y = ~pct_havo_plus,
      name = group,
      marker = list(
        color = group_colours[[group]],
        size = 5,
        opacity = 0.15
      ),
      hoverinfo = "skip",
      visible = FALSE,
      showlegend = FALSE,
      legendgroup = group
    )
}

# Overall relationship, visible initially.
interactive_plot <- interactive_plot |>
  add_lines(
    data = overall_data,
    x = ~pct_test,
    y = ~fit,
    line = list(width = 3),
    hovertemplate = paste(
      "Test performance: %{x:.1f}%",
      "<br>Predicted HAVO+ advice: %{y:.1f}%",
      "<extra></extra>"
    ),
    showlegend = FALSE
  )

# Disadvantage-specific prediction lines, hidden initially.
for (group in levels(prediction_interactive$disadvantage)) {
  line_data <- prediction_interactive |>
    filter(disadvantage == group)

  interactive_plot <- interactive_plot |>
    add_lines(
      data = line_data,
      x = ~pct_test,
      y = ~fit,
      name = group,
      text = ~hover_text,
      hoverinfo = "text",
      line = list(color = group_colours[[group]], width = 4),
      visible = FALSE,
      showlegend = TRUE,
      legendgroup = group
    )
}

interactive_plot <- interactive_plot |>
  layout(
    margin = list(l = 75, r = 25, t = 160, b = 70),
    annotations = list(
      list(
        x = 0,
        y = 1.20,
        xref = "paper",
        yref = "paper",
        text = "<b>The advice gap widens as test performance rises</b>",
        showarrow = FALSE,
        xanchor = "left",
        align = "left",
        font = list(size = 18)
      ),
      list(
        x = 0,
        y = 1.155,
        xref = "paper",
        yref = "paper",
        text = paste0(
          "Compare the overall relationship with predictions at different ",
          "levels of school disadvantage."
        ),
        showarrow = FALSE,
        xanchor = "left",
        align = "left",
        font = list(size = 12)
      )
    ),
    xaxis = list(
      title = "% reaching the target test level",
      range = c(20, 85),
      automargin = TRUE
    ),
    yaxis = list(
      title = "% advised HAVO or higher",
      range = c(0, 70),
      automargin = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      y = 0.14,
      xanchor = "center",
      yanchor = "middle"
    ),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0,
        y = 1.03,
        xanchor = "left",
        yanchor = "bottom",
        showactive = TRUE,
        buttons = list(
          list(
            label = "Overall",
            method = "update",
            args = list(
              list(
                visible = c(
                  TRUE,
                  FALSE, FALSE, FALSE,
                  TRUE,
                  FALSE, FALSE, FALSE
                )
              ),
              list(showlegend = FALSE)
            )
          ),
          list(
            label = "Show school disadvantage",
            method = "update",
            args = list(
              list(
                visible = c(
                  FALSE,
                  TRUE, TRUE, TRUE,
                  FALSE,
                  TRUE, TRUE, TRUE
                )
              ),
              list(showlegend = TRUE)
            )
          )
        )
      )
    )
  )

interactive_plot
```

## Current interpretation

The plot is designed to show that schools with similar test performance do not have the same predicted HAVO-or-higher advice across levels of school disadvantage. The difference is small at lower test-performance levels and becomes larger as test performance rises.

This interactive version is the current working draft of the final visualization. Formatting, uncertainty display, and possible provider-related refinements can be considered later without changing the core analysis above.
