---
title: "Advice Outcome Analysis"
author: "Group 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

source("scripts/00-packages.R")

source(here("scripts", "01-get-data.R"))
```

## Purpose

This notebook keeps the main analytical decisions for RQ3 in one place:

> **Do schools with similar test results give similar secondary-school advice?**

The final visualization should show whether school disadvantage (`schoolweging`)
changes the relationship between school-level test performance and secondary
school advice.

This file is deliberately exploratory and compact. It records:

1. why **HAVO+** was selected as the main advice outcome;
2. why **K+** is used as a robustness threshold;
3. the first moderation results for both outcomes;
4. what those results imply for the final visualization.

## 1. Advice outcomes

### Data preparation

The advice data contain privacy-suppressed counts written as `"<5"`. For the
current analysis these cells are replaced by **2.5**, the midpoint of 1--4.
Genuine missing values remain `NA`.

Split-track advice such as `HAVO_VWO` is assigned **0.5 to either side of a
threshold**. This is an analytical midpoint convention, not a rule specified by
the source data.

```{r advice-prep}
names(schooladviezen) <- gsub("\\u00a0", "", names(schooladviezen))

advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)

advice <- schooladviezen |>
  mutate(
    across(
      all_of(advice_cols),
      ~ if_else(.x == "<5", 2.5, as.numeric(.x))
    )
  ) |>
  mutate(n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE)) |>
  filter(n_advice > 0) |>
  mutate(
    pct_vwo = 100 * VWO / n_advice,
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice,
    pct_gt_plus = 100 * (
      0.5 * VMBO_GT_HAVO + VMBO_GT + HAVO + 0.5 * HAVO_VWO + VWO
    ) / n_advice,
    pct_k_plus = 100 * (
      0.5 * VMBO_B_K + VMBO_K + 0.5 * VMBO_K_GT + VMBO_GT +
        0.5 * VMBO_GT_HAVO + HAVO + 0.5 * HAVO_VWO + VWO
    ) / n_advice,
    pct_vmbo = 100 * (
      VMBO_B + VMBO_B_K + VMBO_K + VMBO_K_GT + VMBO_GT +
        0.5 * VMBO_GT_HAVO
    ) / n_advice,
    pct_pro_vso = 100 * (PRO + VSO) / n_advice
  )
```

### Suppression burden

The useful question is not simply whether a school contains a suppressed cell,
because almost all schools do. Instead, we compare how much of each reconstructed
outcome comes from cells imputed as 2.5.

```{r suppression}
raw_advice <- schooladviezen
names(raw_advice) <- gsub("\\u00a0", "", names(raw_advice))

suppression_data <- raw_advice |>
  mutate(
    across(
      all_of(advice_cols),
      ~ if_else(.x == "<5", 2.5, 0),
      .names = "sup_{.col}"
    ),
    across(
      all_of(advice_cols),
      ~ if_else(.x == "<5", 2.5, as.numeric(.x))
    )
  ) |>
  mutate(n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE)) |>
  filter(n_advice > 0) |>
  mutate(
    sup_vwo = sup_VWO,
    sup_havo_plus = sup_HAVO + 0.5 * sup_HAVO_VWO + sup_VWO,
    sup_gt_plus = 0.5 * sup_VMBO_GT_HAVO + sup_VMBO_GT + sup_HAVO +
      0.5 * sup_HAVO_VWO + sup_VWO,
    sup_k_plus = 0.5 * sup_VMBO_B_K + sup_VMBO_K + 0.5 * sup_VMBO_K_GT +
      sup_VMBO_GT + 0.5 * sup_VMBO_GT_HAVO + sup_HAVO +
      0.5 * sup_HAVO_VWO + sup_VWO,
    sup_vmbo = sup_VMBO_B + sup_VMBO_B_K + sup_VMBO_K + sup_VMBO_K_GT +
      sup_VMBO_GT + 0.5 * sup_VMBO_GT_HAVO,
    sup_pro_vso = sup_PRO + sup_VSO,
    n_vwo = VWO,
    n_havo_plus = HAVO + 0.5 * HAVO_VWO + VWO,
    n_gt_plus = 0.5 * VMBO_GT_HAVO + VMBO_GT + HAVO +
      0.5 * HAVO_VWO + VWO,
    n_k_plus = 0.5 * VMBO_B_K + VMBO_K + 0.5 * VMBO_K_GT + VMBO_GT +
      0.5 * VMBO_GT_HAVO + HAVO + 0.5 * HAVO_VWO + VWO,
    n_vmbo = VMBO_B + VMBO_B_K + VMBO_K + VMBO_K_GT + VMBO_GT +
      0.5 * VMBO_GT_HAVO,
    n_pro_vso = PRO + VSO
  )

suppression_summary <- tibble(
  outcome = c("VWO", "HAVO+", "GT+", "K+", "VMBO", "PRO + VSO"),
  imputed_count = c(
    sum(suppression_data$sup_vwo, na.rm = TRUE),
    sum(suppression_data$sup_havo_plus, na.rm = TRUE),
    sum(suppression_data$sup_gt_plus, na.rm = TRUE),
    sum(suppression_data$sup_k_plus, na.rm = TRUE),
    sum(suppression_data$sup_vmbo, na.rm = TRUE),
    sum(suppression_data$sup_pro_vso, na.rm = TRUE)
  ),
  reconstructed_count = c(
    sum(suppression_data$n_vwo, na.rm = TRUE),
    sum(suppression_data$n_havo_plus, na.rm = TRUE),
    sum(suppression_data$n_gt_plus, na.rm = TRUE),
    sum(suppression_data$n_k_plus, na.rm = TRUE),
    sum(suppression_data$n_vmbo, na.rm = TRUE),
    sum(suppression_data$n_pro_vso, na.rm = TRUE)
  )
) |>
  mutate(imputed_share = 100 * imputed_count / reconstructed_count)

suppression_summary
```

The reproduced results are:

| Outcome | Imputed contribution |
|---|---:|
| VWO | 19.6% |
| HAVO+ | 26.9% |
| GT+ | 30.3% |
| K+ | 37.2% |
| VMBO | 53.4% |
| PRO + VSO | 53.9% |

**Interpretation:** HAVO+ is less dependent on imputed suppressed cells than the
broader lower-track measures. K+ has more imputation than HAVO+, but remains
usable as a robustness threshold.

### How similar are the outcomes?

```{r outcome-correlations}
outcome_cols <- c(
  "pct_vwo", "pct_havo_plus", "pct_gt_plus", "pct_k_plus",
  "pct_vmbo", "pct_pro_vso"
)

cor_with_havo <- advice |>
  summarise(
    across(
      all_of(outcome_cols),
      ~ cor(.x, pct_havo_plus, use = "complete.obs")
    )
  )

cor_with_havo
```

The reproduced correlations with HAVO+ are:

| Outcome | Correlation with HAVO+ |
|---|---:|
| VWO | 0.831 |
| HAVO+ | 1.000 |
| GT+ | 0.855 |
| K+ | 0.715 |
| VMBO | -0.777 |
| PRO + VSO | -0.472 |

**Interpretation:**

- **HAVO+** is the preferred main outcome: it is easy to interpret and has a
  relatively lower suppression burden.
- **VWO** and **GT+** are highly redundant with HAVO+.
- **All VMBO** mainly restates the HAVO+/VMBO divide from the opposite direction.
- **K+** is the most useful robustness threshold because it moves the cutoff
  substantially lower in the ordered track hierarchy.
- **PRO and VSO** are substantively different pathways and are not used as the
  main robustness outcome.

### Outcome decision

The next analysis therefore uses:

- **Primary outcome:** HAVO or higher (`pct_havo_plus`)
- **Robustness outcome:** VMBO-K or higher (`pct_k_plus`)

The K+ analysis tests whether the disadvantage pattern is broad across the
advice hierarchy or concentrated around the HAVO/VMBO boundary.

## 2. Test-performance measure

The current assignment analysis defines test performance as the average of:

- the percentage reaching **maths 1S**, and
- the percentage reaching **reading 2F**.

Maths and reading showed similar relationships with advice in the exploratory
analysis, supporting their use as a combined school-level test measure.

```{r test-performance}
to_num <- function(x) {
  if_else(x == "<5", 2.5, as.numeric(x))
}

ref <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  )
```

**Interpretation:** the test measure is provider-independent because it uses
standardized reference-level attainment rather than raw provider scores.

## 3. First moderation results

The first moderation analysis compared:

```r
advice ~ test_performance * schoolweging
```

for HAVO+ and K+. `schoolweging` was kept continuous.

The results below record the output from Session 2. The next refinement step
centers the predictors and converts these results into more interpretable
predicted values; centering does not change the fitted relationship.

### HAVO+

- Correlation with test performance: **0.65**
- Correlation with `schoolweging`: **-0.60**
- Full interaction model: **R² = 0.53**
- Interaction: **-0.033**

The predicted p90-minus-p10 `schoolweging` gap was:

| Test performance | Higher- minus lower-disadvantage gap |
|---:|---:|
| 25 | -3.5 pp |
| 40 | -8.3 pp |
| 55 | -13.1 pp |
| 70 | -18.0 pp |

The estimated test-performance slope was approximately:

- **0.63** at lower disadvantage
- **0.31** at higher disadvantage

**Interpretation:** the HAVO+ disadvantage gap is not constant. It becomes much
larger as school-level test performance increases. Among lower-performing
schools the predicted difference is modest, while among higher-performing
schools the predicted difference in HAVO/VWO advice is substantially larger.

This means the central HAVO+ result is mainly an **interaction / slope
difference**, rather than a simple constant downward shift.

### K+

- Correlation with test performance: **0.35**
- Correlation with `schoolweging`: **-0.37**
- Full interaction model: **R² = 0.17**
- Interaction: **+0.006**, which is small

The predicted p90-minus-p10 `schoolweging` gap was:

| Test performance | Higher- minus lower-disadvantage gap |
|---:|---:|
| 25 | -8.7 pp |
| 40 | -7.8 pp |
| 55 | -7.0 pp |
| 70 | -6.1 pp |

**Interpretation:** K+ shows a disadvantage difference, but it is approximately
constant across test performance. Unlike HAVO+, the K+ result is mainly a
**level shift**, not a slope difference.

## 4. What the two outcomes tell us

The two outcomes do **not** show exactly the same pattern.

- For **HAVO+**, the disadvantage gap grows strongly as test performance rises.
- For **K+**, the gap is much flatter across the test-performance range.

This suggests that disadvantage is associated with lower advice more broadly,
but the especially strong widening gap at higher performance is concentrated
around reaching the **HAVO/VWO** part of the hierarchy.

This strengthens HAVO+ as the main outcome for the final visualization rather
than weakening it: HAVO+ captures the part of the advice distribution where the
relationship with disadvantage changes most clearly with performance.

## 5. Implication for the final visualization

The first moderation result favors a **predicted-lines visualization** over a
single residual-gap plot.

The likely main plot is:

- **x-axis:** school-level test performance
- **y-axis:** percentage advised HAVO or higher
- predicted lines for representative values of continuous `schoolweging`

The key visual message should be the **divergence of the predicted lines at
higher test performance**, rather than the interaction coefficient itself.

The final display should use representative disadvantage levels such as the
25th, 50th, and 75th percentiles rather than categorizing `schoolweging` in the
statistical model.

## 6. Simple checks for the first visualization draft

At this stage, centering the predictors is not necessary. Centering changes the
interpretation of the regression coefficients, but it does **not** change the
fitted relationship or the predicted lines. For the first visualization draft,
the useful quantities are therefore the model predictions at representative
levels of `schoolweging`.

### Build one modelling dataset

```{r model-data}
adv_raw <- schooladviezen
names(adv_raw) <- gsub("\\u00a0", "", names(adv_raw))

adv_ids <- adv_raw[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

adv_model <- adv_raw |>
  mutate(
    inst = adv_ids[[1]],
    vest = adv_ids[[2]],
    across(
      all_of(advice_cols),
      ~ if_else(.x == "<5", 2.5, as.numeric(.x))
    )
  ) |>
  mutate(
    n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice
  ) |>
  filter(n_advice > 0) |>
  select(inst, vest, pct_havo_plus)

ref_model <- referentieniveaus |>
  mutate(
    inst = ref_ids[[1]],
    vest = ref_ids[[2]],
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  ) |>
  select(inst, vest, pct_test)

sw_model <- schoolweging |>
  transmute(
    inst = sub("\\|.*$", "", OVT),
    schoolweging
  ) |>
  filter(!is.na(schoolweging)) |>
  distinct(inst, .keep_all = TRUE)

model_data <- adv_model |>
  inner_join(ref_model, by = c("inst", "vest")) |>
  inner_join(sw_model, by = "inst") |>
  filter(!is.na(pct_test), !is.na(pct_havo_plus))

nrow(model_data)
```

### Predicted HAVO+ lines at representative disadvantage levels

The statistical model keeps `schoolweging` continuous. The 25th, 50th, and
75th percentiles are used only to display representative predicted lines.

```{r havo-predictions}
model_havo <- lm(
  pct_havo_plus ~ pct_test * schoolweging,
  data = model_data
)

sw_values <- quantile(
  model_data$schoolweging,
  probs = c(0.25, 0.50, 0.75),
  na.rm = TRUE
)

sw_values

prediction_data <- expand_grid(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  ),
  schoolweging = as.numeric(sw_values)
) |>
  mutate(
    disadvantage = factor(
      schoolweging,
      levels = as.numeric(sw_values),
      labels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    )
  )

prediction_data <- bind_cols(
  prediction_data,
  as.data.frame(
    predict(model_havo, newdata = prediction_data, interval = "confidence")
  )
)

summary(model_havo)$r.squared
coef(model_havo)
```

For the final visualization, the important result is not the raw regression
coefficient but how far apart the predicted lines are at the same test result.

```{r predicted-gaps}
gap_data <- expand_grid(
  pct_test = c(25, 40, 55, 70),
  schoolweging = c(sw_values[[1]], sw_values[[3]])
) |>
  mutate(
    predicted = predict(model_havo, newdata = pick(everything()))
  ) |>
  arrange(pct_test, schoolweging) |>
  group_by(pct_test) |>
  summarise(
    low_disadvantage = first(predicted),
    high_disadvantage = last(predicted),
    gap = high_disadvantage - low_disadvantage,
    .groups = "drop"
  )

gap_data
```

**Interpretation:** if the gap becomes increasingly negative as test performance
rises, the first visualization should emphasize the divergence between the
lower- and higher-disadvantage predicted lines.

### Quick sensitivity check

The goal here is not to create a second full analysis. We only check whether the
main interaction changes materially under plausible coding choices.

The function below refits the same HAVO+ model while changing:

- the replacement for `"<5"`; and
- the share of `HAVO_VWO` counted above the HAVO threshold.

It returns the interaction coefficient plus the predicted p75-minus-p25
disadvantage gap at test-performance values 40 and 70.

```{r sensitivity-function}
run_check <- function(suppressed = 2.5, split = 0.5) {
  num <- function(x) {
    if_else(x == "<5", suppressed, as.numeric(x))
  }

  adv_check <- adv_raw |>
    mutate(
      inst = adv_ids[[1]],
      vest = adv_ids[[2]],
      across(all_of(advice_cols), num)
    ) |>
    mutate(
      n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
      pct_havo_plus = 100 * (HAVO + split * HAVO_VWO + VWO) / n_advice
    ) |>
    filter(n_advice > 0) |>
    select(inst, vest, pct_havo_plus)

  ref_check <- referentieniveaus |>
    mutate(
      inst = ref_ids[[1]],
      vest = ref_ids[[2]],
      tot_m = num(REKENEN_LAGER1F) + num(REKENEN_1F) +
        num(REKENEN_1S) + num(REKENEN_2F),
      tot_r = num(LV_LAGER1F) + num(LV_1F) + num(LV_2F),
      pct_m = 100 * num(REKENEN_1S) / tot_m,
      pct_r = 100 * num(LV_2F) / tot_r,
      pct_test = (pct_m + pct_r) / 2
    ) |>
    select(inst, vest, pct_test)

  dat <- adv_check |>
    inner_join(ref_check, by = c("inst", "vest")) |>
    inner_join(sw_model, by = "inst") |>
    filter(!is.na(pct_test), !is.na(pct_havo_plus))

  fit <- lm(pct_havo_plus ~ pct_test * schoolweging, data = dat)
  sw <- quantile(dat$schoolweging, c(0.25, 0.75), na.rm = TRUE)

  pred <- expand_grid(
    pct_test = c(40, 70),
    schoolweging = as.numeric(sw)
  ) |>
    mutate(predicted = predict(fit, newdata = pick(everything()))) |>
    select(pct_test, schoolweging, predicted) |>
    pivot_wider(names_from = schoolweging, values_from = predicted)

  tibble(
    suppressed = suppressed,
    split = split,
    interaction = coef(fit)[["pct_test:schoolweging"]],
    gap_test_40 = pred[[3]][pred$pct_test == 40] -
      pred[[2]][pred$pct_test == 40],
    gap_test_70 = pred[[3]][pred$pct_test == 70] -
      pred[[2]][pred$pct_test == 70]
  )
}
```

First vary only the privacy-suppression assumption:

```{r sensitivity-suppression}
suppression_check <- bind_rows(
  run_check(suppressed = 1, split = 0.5),
  run_check(suppressed = 2.5, split = 0.5),
  run_check(suppressed = 4, split = 0.5)
)

suppression_check
```

Then vary only the treatment of `HAVO_VWO`:

```{r sensitivity-split}
split_check <- bind_rows(
  run_check(suppressed = 2.5, split = 0),
  run_check(suppressed = 2.5, split = 0.5),
  run_check(suppressed = 2.5, split = 1)
)

split_check
```

**Interpretation rule:** the main conclusion is robust enough for Draft 1 if:

- the interaction remains negative in all scenarios; and
- the disadvantage gap at test performance 70 remains clearly larger than the
  gap at test performance 40.

Exact coefficients do not need to be identical.

## 7. Ready for Draft 1

If the representative predictions reproduce the widening HAVO+ gap and the two
quick sensitivity checks preserve that pattern, the analysis is sufficiently
stable to begin the first final-visualization draft.

The first draft can then use:

- **x-axis:** school-level test performance;
- **y-axis:** percentage advised HAVO or higher;
- **three predicted lines:** p25, p50, and p75 `schoolweging`;
- a direct annotation at the higher-performance end showing the predicted
  percentage-point difference between lower- and higher-disadvantage schools.

The provider analysis can follow **after Draft 1**. Its purpose will be to test
whether provider materially changes the pattern and therefore whether the draft
needs an additional provider layer or qualification.
```{r first visualization draft}
# First draft of the main visualization
fig_draft <- prediction_data |>
  ggplot(aes(x = pct_test, y = fit, colour = disadvantage)) +
  geom_point(
    data = model_data,
    aes(x = pct_test, y = pct_havo_plus),
    inherit.aes = FALSE,
    alpha = 0.08,
    size = 0.8,
    colour = "grey40"
  ) +
  geom_line(linewidth = 1.3) +
  scale_x_continuous(
    breaks = seq(0, 100, 10),
    limits = c(20, 80)
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, 10),
    limits = c(0, 70)
  ) +
  labs(
    x = "% reaching the target test level",
    y = "% advised HAVO or higher",
    colour = "School disadvantage"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

fig_draft

ggsave(
  "figures/final-viz-draft-1.png",
  fig_draft,
  width = 8,
  height = 6,
  dpi = 300
)
```