---
title: "DV Assignment 2: [Your Title Here]"
author: "Group 1: Matheus Nacif, Ilinca Anton, Isabella Braune"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

RQ3. Do schools with similar test results give similar secondary-school advice?

Level: Inference

Target audience: secondary schools, or policymakers.

Instruction: At the end of primary school, every student receives a school advice for a secondary-school track (PRO, VMBO, HAVO, VWO, see the codebook). The advice is based on the teacher’s judgement and on the doorstroomtoets. In this RQ, you will investigate how closely a school’s advice follows its students’ test results, and whether this differs between schools with a more or less disadvantaged student population.

Relate a measure of the school’s advice (e.g. pct_havo_plus) to a measure of its test results (pct_maths_1s and/or pct_reading_2f), and test whether schoolweging moderates this relationship. Do two schools with the same results give the same advice, regardless of who their students are?

*What are you investigating, and why does it matter? Who is the target audience for your final visualization, and how does that shape your choices?*

## Exploratory process

*Describe how you explored the data before settling on your final visualization. Include at least one exploratory plot below, and a short note on what it told you (or didn't).*

```{r exploratory-plot}
```

## Final visualization

*Your final, explanatory visualization goes below. Include a short write-up: what does it communicate, why is this the best way to communicate it, and how robust is the finding?*

```{r final-plot}
```

## Conclusion

*What can you conclude from your visualization? What are the limitations of this analysis?*

---

**Before submitting:** run the lintr check on this file, and re-knit from a clean environment (`rm(list = ls())`, already done above) to make sure everything reproduces from scratch.
