---
title: "DV Assignment 2: [Your Title Here]"
author: "Group 1: Matheus Nacif, Ilinca Anton, Isabella Braune"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source(here::here("scripts", "00-packages.R"))

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here::here("scripts", "01-get-data.R"))

# loads the custom ggplot and plotly themes for this project 
# see scripts/02-custom_themes.R for the available functions. 
# the theme should only be edited within this file 
source(here::here("scripts", "02-custom_themes.R"))
```

## Research question

RQ3. Do schools with similar test results give similar secondary-school advice?

Level: Inference

Target audience: secondary schools, or policymakers.

Instruction: At the end of primary school, every student receives a school advice for a secondary-school track (PRO, VMBO, HAVO, VWO, see the codebook). The advice is based on the teacher’s judgement and on the doorstroomtoets. In this RQ, you will investigate how closely a school’s advice follows its students’ test results, and whether this differs between schools with a more or less disadvantaged student population.

Relate a measure of the school’s advice (e.g. pct_havo_plus) to a measure of its test results (pct_maths_1s and/or pct_reading_2f), and test whether schoolweging moderates this relationship. Do two schools with the same results give the same advice, regardless of who their students are?

### What are we investigating, why does it matter?

Every primary-school pupil receives a secondary-track advice at the end of group 8 (PRO, VMBO, HAVO, or VWO) that sets their entire secondary trajectory. This advice is based on the teacher's judgement and on the doorstroomtoets, a standardized test designed to be comparable across schools. We ask whether, holding test performance constant, a school's advice still depends on how socio-economically disadvantaged its student body is (`schoolweging`). Concretely, we relate each school's share of students advised to HAVO/VWO to its share reaching the reference level in maths and/or reading, and test whether `schoolweging` shifts that relationship. It matters because if disadvantaged schools systematically assign lower tracks to students with the same demonstrated performance, then a standardized test is failing its purpose and students are being pushed down the track on the basis of context rather than ability, a live fairness concern in the Dutch debate over transfer advice.

### Who is the target Audience, how does that shape the choices?

Our primary audience is **policymakers and the general public** (a newspaper-graphic reader), with school boards as a secondary audience. Since this reader is not a statistician, the visualization must carry *one* legible conclusion (advice tracks test results, and disadvantage shifts it by this much) expressed in concrete, relatable units (e.g. percentage points of students advised upward) rather than coefficients or p-values. That pushes us toward a small number of clear encodings (a fit for advantaged vs. disadvantaged schools, or a gap/residual line) and toward a single highlighted pattern, rather than a dense multivariate plot.

#### To Discuss:

We need to check how we want to summarise the variables. Two candidates: (1) `pct_havo_plus` — the share of students advised to HAVO or VWO, a single binary-style metric that is easy for a general audience to read; or (2) a numeric track scale (PRO = 1, VMBO = 2, HAVO = 3, VWO = 6) averaged per school, which uses the full range of the advice and does not collapse the lower tracks (PRO, SBO, VMBO) into one bucket. The choice matters: the binary is simpler to communicate, while the numeric scale is more faithful to how much of a gap a track move represents. We can pick between them once the exploratory plots make clear which measure carries the pattern we want to show.

## Exploratory process

#### Figure 1: Distribution of school weighting across schools

```{r exploratory-plot-1, fig.height = 4, fig.width = 6, fig.align = "center"}
# Drop the 119 empty locations (0 students, no schoolweging)
sw <- schoolweging |>
  filter(!is.na(schoolweging))

sw |>
  ggplot(aes(x = schoolweging)) +
  geom_histogram(bins = 40, fill = "#2878CC", color = "white", alpha = 0.7) +
  geom_density(fill = "#2878CC", alpha = 0.08) +
  geom_vline(
    xintercept = median(sw$schoolweging),
    linetype = "dashed", color = "grey30"
  ) +
  annotate(
    "text",
    x = median(sw$schoolweging) + 0.5,
    y = max(table(cut(sw$schoolweging, breaks = 40))) * 0.85,
    label = paste0("median = ", round(median(sw$schoolweging), 1)),
    hjust = 0, size = 3.5, color = "grey30"
  ) +
  scale_x_continuous(limits = c(18, 43)) +
  labs(
    x = "School weighting (higher = more disadvantaged)",
    y = "Number of schools"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The distribution of `schoolweging` is roughly bell-shaped, centred around 29–30, with the bulk of schools falling between 22 and 37. There is no natural bimodal split — the spread is continuous. For the moderation analysis we'll need to pick a threshold (e.g. median split at \~29.5, or tertiles) to define "low" vs "high" disadvantage groups; the plot makes clear that no single school is an outlier, and that both tails have enough mass to compare.

#### Figure 2: Share of students advised HAVO or higher, by share reaching the reference level

```{r exploratory-plot-2, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}
# the count columns are character ("<5" privacy suppression) -> numeric first
adv_cols <- c("PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT", "VMBO_GT",
              "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO", "ADVIES_NIET_MOGELIJK")
to_num <- function(x) {
  v <- suppressWarnings(as.numeric(x))
  v[is.na(v)] <- 2.5
  v
}

# pull the two id columns out before mutating (their headers carry a
# hidden character and their position differs between the two files)
adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

adv <- schooladviezen |>
  mutate(across(all_of(adv_cols), to_num)) |>
  mutate(
    n_advice = rowSums(across(all_of(adv_cols))),
    pct_havo_plus = 100 * (HAVO + HAVO_VWO + VWO) / n_advice
  )

# proportion of pupils reaching the reference level, averaged across
# maths (1S) and reading (2F) — removes school size as a confound
ref <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_ref = (pct_m + pct_r) / 2
  )


plot_data <- data.frame(inst = adv_ids[[1]], vest = adv_ids[[2]],
                        n_advice = adv$n_advice,
                        pct_havo_plus = adv$pct_havo_plus) |>
  inner_join(
    data.frame(inst = ref_ids[[1]], vest = ref_ids[[2]], pct_ref = ref$pct_ref),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(pct_ref))

plot_data |>
  ggplot(aes(x = pct_ref, y = pct_havo_plus)) +
  geom_point(alpha = 0.3, size = 1, color = "#2878CC") +
  geom_smooth(method = "lm", color = "#EB6734", se = TRUE) +
  labs(
    x = "% of students reaching the reference level (maths 1S + reading 2F)",
    y = "% of students advised to HAVO or higher"
  ) +
  theme_minimal()
```

Figure 2 shows that across the 5,921 schools, the relationship between test result and advice is clearly positive (correlation 0.69): schools whose students reach the reference level in maths and reading advise a larger share of them to HAVO or VWO. On the whole, advice tracks demonstrated performance.

However, our test measure averages two different subjects. Maths and reading measure different skills, and their target levels can differ in difficulty. Averaging them assumes that advice responds to both subjects in the same way.

If advice actually depends more on one subject than the other, averaging them weakens this claim. This would in turn make it unclear what "schools with the same results" means, and that comparison is the core of our research question. We therefore looked at the two subjects separately in figure 3, relating each school's share of VWO advice, the highest track, to its share of students reaching the maths target (1S) and the reading target (2F).

#### Figure 3:

```{r exploratory-plot-3, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}

# share of students reaching the target level, per subject separately
ref_subjects <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) + to_num(REKENEN_1S),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    maths = 100 * to_num(REKENEN_1S) / tot_m,
    reading = 100 * to_num(LV_2F) / tot_r
  )

# one row per school per subject, with readable subject names for the legend
subject_data <- data.frame(
  inst = adv_ids[[1]], vest = adv_ids[[2]],
  n_advice = adv$n_advice,
  pct_vwo = 100 * adv$VWO / adv$n_advice
) |>
  inner_join(
    data.frame(
      inst = ref_ids[[1]], vest = ref_ids[[2]],
      maths = ref_subjects$maths, reading = ref_subjects$reading
    ),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(maths), !is.na(reading)) |>
  pivot_longer(
    c(maths, reading),
    names_to = "subject", values_to = "pct_target"
  ) |>
  mutate(
    label = if_else(subject == "maths", "Maths (1S)", "Reading (2F)")
  ) # Make readable labels for the legend


subject_data |>
  ggplot(
    aes(x = pct_target, y = pct_vwo, colour = label, fill = label)
  ) +
  # shape 21 is a filled circle, so the points use the fill colours
  geom_point(shape = 21, stroke = 0, alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm", formula = y ~ x, se = TRUE,
    alpha = 0.25, linewidth = 1.3
  ) +
  # light colours for the points, darker shades of the same for the lines
  scale_fill_manual(values = c("#009E73", "#CC79A7"), name = NULL) +
  scale_colour_manual(values = c("#00553E", "#7A2E5C"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 25)) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100), expand = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  labs(
    x = "% of students reaching the target level",
    y = "% of students advised VWO"
  ) +
  theme_minimal() +
  theme(legend.position = "top")


# Looking at the correlation:
subject_data |>
  group_by(subject) |>
  summarise(r = cor(pct_target, pct_vwo))
```

Looking at Figure 3, we observe that advice is moderately related to both: slightly more strongly to maths (r = 0.55) than to reading (r = 0.49). Since the difference is small, advice does not appear to depend mainly on one subject. This supports using the average of both subjects as our measure of test results. Because the data is at school level, this describes how schools' advice relates to their overall results, not how individual students are advised.

## Conclusion

School weighting is spread continuously across schools, roughly bell-shaped around 29–30, with enough schools at both ends to compare more and less disadvantaged populations (Figure 1). Schools with better test results clearly advise more students to HAVO or higher (r = 0.69; Figure 2), so on the whole, advice follows demonstrated performance. Looking at the two subjects separately, advice is related to maths (r = 0.55) and reading (r = 0.49) about equally, which supports using the average of both as our measure of test results (Figure 3).

At the same time, figure 2 and 3 both show that schools with similar test results still differ considerably in how many students they advise to a higher track. The question for the next phase is whether schoolweging explains part of that difference. At the same test results, does a more disadvantaged school advise fewer of its students to a higher track?

## Moderation analysis

The exploratory analysis above supports using the average of maths 1S and reading 2F as the measure of school-level test performance. We now test whether `schoolweging` changes the relationship between test performance and HAVO-or-higher advice.

For this analysis:

- privacy-suppressed counts (`"<5"`) are replaced by **2.5**;
- split-track advice (`HAVO_VWO`) contributes **0.5** to the HAVO-or-higher outcome;
- `schoolweging` remains continuous in the model;
- the 25th, 50th, and 75th percentiles of `schoolweging` are used only to visualize representative levels of disadvantage.

### Prepare the analysis data

```{r prepare-moderation-data}
names(schooladviezen) <- gsub("\\u00a0", "", names(schooladviezen))

advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)

to_num <- function(x) {
  if_else(x == "<5", 2.5, suppressWarnings(as.numeric(x)))
}

adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

advice_data <- schooladviezen |>
  mutate(
    inst = adv_ids[[1]],
    vest = adv_ids[[2]],
    across(all_of(advice_cols), to_num),
    n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice
  ) |>
  filter(n_advice > 0) |>
  select(inst, vest, pct_havo_plus)

reference_data <- referentieniveaus |>
  mutate(
    inst = ref_ids[[1]],
    vest = ref_ids[[2]],
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  ) |>
  select(inst, vest, pct_test)

weighting_data <- schoolweging |>
  transmute(
    inst = sub("\\|.*$", "", OVT),
    schoolweging
  ) |>
  filter(!is.na(schoolweging)) |>
  distinct(inst, .keep_all = TRUE)

model_data <- advice_data |>
  inner_join(reference_data, by = c("inst", "vest")) |>
  inner_join(weighting_data, by = "inst") |>
  filter(!is.na(pct_test), !is.na(pct_havo_plus))

nrow(model_data)
```

The analysis is performed at the **school level**. The outcome is the percentage receiving HAVO-or-higher advice, while test performance is based on standardized reference-level attainment rather than provider-specific raw scores.

### Model school disadvantage as a moderator

The interaction model tests whether the relationship between test performance and HAVO-or-higher advice changes with school disadvantage.

```{r moderation-model}
model_havo <- lm(
  pct_havo_plus ~ pct_test * schoolweging,
  data = model_data
)

model_results <- tibble(
  r_squared = summary(model_havo)$r.squared,
  interaction = coef(model_havo)[["pct_test:schoolweging"]]
)

model_results
```

The interaction is negative, meaning that the positive relationship between test performance and HAVO-or-higher advice becomes weaker as school disadvantage increases. In other words, the difference between lower- and higher-disadvantage schools becomes larger at higher levels of test performance.

#### Representative predictions

The model keeps `schoolweging` continuous. The three lines used in the visualization represent its 25th, 50th, and 75th percentiles.

```{r moderation-predictions}
sw_values <- quantile(
  model_data$schoolweging,
  probs = c(0.25, 0.50, 0.75),
  na.rm = TRUE
)

prediction_data <- expand_grid(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  ),
  schoolweging = as.numeric(sw_values)
) |>
  mutate(
    disadvantage = factor(
      schoolweging,
      levels = as.numeric(sw_values),
      labels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    ),
    fit = predict(model_havo, newdata = pick(pct_test, schoolweging))
  )
```

To make the widening gap easier to interpret, we compare the lower- and higher-disadvantage predictions at several test-performance values.

```{r moderation-gap-check}
gap_check <- expand_grid(
  pct_test = c(25, 40, 55, 70),
  schoolweging = c(sw_values[[1]], sw_values[[3]])
) |>
  mutate(predicted = predict(model_havo, newdata = pick(everything()))) |>
  arrange(pct_test, schoolweging) |>
  group_by(pct_test) |>
  summarise(
    lower_disadvantage = first(predicted),
    higher_disadvantage = last(predicted),
    gap = higher_disadvantage - lower_disadvantage,
    .groups = "drop"
  )

gap_check
```

The predicted difference is small at lower test-performance levels and widens as test performance rises. At 70% test performance, the predicted HAVO-or-higher advice gap between the 25th and 75th percentiles of school disadvantage is about **8 percentage points**.

### Interactive visualization

The interactive plot has two states:

1. **Overall:** observed schools and the overall relationship between test performance and HAVO-or-higher advice.
2. **Show school disadvantage:** schools are coloured by broad disadvantage group and the overall line is replaced by predictions at the 25th, 50th, and 75th percentiles of `schoolweging`.

The broad groups used to colour the points are only a visual aid. The model itself continues to use continuous `schoolweging`.

```{r interactive-data}
model_data <- model_data |>
  mutate(
    disadvantage_group = case_when(
      schoolweging <= sw_values[[1]] ~ "Lower disadvantage",
      schoolweging >= sw_values[[3]] ~ "Higher disadvantage",
      TRUE ~ "Typical"
    ),
    disadvantage_group = factor(
      disadvantage_group,
      levels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    )
  )

high_reference <- prediction_data |>
  filter(disadvantage == "Higher disadvantage") |>
  select(pct_test, high_fit = fit)

prediction_interactive <- prediction_data |>
  left_join(high_reference, by = "pct_test") |>
  mutate(
    difference_high = fit - high_fit,
    hover_text = paste0(
      "<b>", disadvantage, "</b>",
      "<br>Test performance: ", round(pct_test, 1), "%",
      "<br>Predicted HAVO+ advice: ", round(fit, 1), "%",
      "<br>Difference from higher-disadvantage school: ",
      if_else(
        abs(difference_high) < 1e-8,
        "reference",
        paste0(
          if_else(difference_high > 0, "+", ""),
          round(difference_high, 1),
          " percentage points"
        )
      )
    )
  )

overall_model <- lm(pct_havo_plus ~ pct_test, data = model_data)

overall_data <- tibble(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  )
) |>
  mutate(fit = predict(overall_model, newdata = pick(everything())))
```

```{r interactive-plot}
# removed the call for group_colours here 

interactive_plot <- plot_ly()

# Default state: neutral school dots.
interactive_plot <- interactive_plot |>
  add_markers(
    data = model_data,
    x = ~pct_test,
    y = ~pct_havo_plus,
    marker = list(color = group_neutral_col, size = 5, opacity = 0.35),
    hoverinfo = "skip",
    showlegend = FALSE
  )

# Disadvantage-coloured dots, hidden initially.
for (group in levels(model_data$disadvantage_group)) {
  group_data <- model_data |>
    filter(disadvantage_group == group)

  interactive_plot <- interactive_plot |>
    add_markers(
      data = group_data,
      x = ~pct_test,
      y = ~pct_havo_plus,
      name = group,
      marker = list(
        color = group_colors[[group]],
        size = 5,
        opacity = 0.15
      ),
      hoverinfo = "skip",
      visible = FALSE,
      showlegend = FALSE,
      legendgroup = group
    )
}

# Overall relationship, visible initially.
interactive_plot <- interactive_plot |>
  add_lines(
    data = overall_data,
    x = ~pct_test,
    y = ~fit,
    line = list(color = group_sec_text,
                width = 3),
    hovertemplate = paste(
      "Test performance: %{x:.1f}%",
      "<br>Predicted HAVO+ advice: %{y:.1f}%",
      "<extra></extra>"
    ),
    showlegend = FALSE
  )

# Disadvantage-specific prediction lines, hidden initially.
for (group in levels(prediction_interactive$disadvantage)) {
  line_data <- prediction_interactive |>
    filter(disadvantage == group)

  interactive_plot <- interactive_plot |>
    add_lines(
      data = line_data,
      x = ~pct_test,
      y = ~fit,
      name = group,
      text = ~hover_text,
      hoverinfo = "text",
      line = list(color = group_colors[[group]], width = 4),
      visible = FALSE,
      showlegend = TRUE,
      legendgroup = group
    )
}

interactive_plot <- interactive_plot |>
  layout(
    margin = list(l = 75, r = 25, t = 125, b = 95),
    
    # place the title and subtitle at a fixed pixel distance above the plot
    # ensures spacing stays the same when the plot is resized 
    annotations = list(
      # --- title --- 
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "left", 
        yanchor = "bottom",
        align = "left", 
        yshift = 88,
        text = "<b>The advice gap widens as test performance rises</b>",
        showarrow = FALSE,
        xanchor = "left",
        align = "left",
        font = list(size = group_px$title)
      ),
      # --- subtitle --- 
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        yanchor = "bottom",
        yshift = 66,
        text = paste0(
          "Compare the overall relationship with predictions at different ",
          "levels of school disadvantage."
        ),
        showarrow = FALSE,
        xanchor = "left",
        align = "left",
        font = list(size = group_px$subtitle,
                    color = group_sec_text)
      )
    ),
    xaxis = list(
      title = "% reaching the target test level",
      range = c(20, 85),
      showline = TRUE, # show the line above the x-axis 
      automargin = TRUE
    ),
    yaxis = list(
      title = "% advised HAVO or higher",
      range = c(-3, 70),
      automargin = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      y = 0.14,
      xanchor = "center",
      yanchor = "middle"
    ),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0,
        y = 1,
        xanchor = "left",
        yanchor = "bottom",
        showactive = TRUE,
        pad = list(b = 10), # adds empty space around the buttons 
        buttons = list(
          list(
            label = "Overall",
            method = "update",
            args = list(
              list(
                visible = c(
                  TRUE,
                  FALSE, FALSE, FALSE,
                  TRUE,
                  FALSE, FALSE, FALSE
                )
              ),
              list(showlegend = FALSE)
            )
          ),
          list(
            label = "Show school disadvantage",
            method = "update",
            args = list(
              list(
                visible = c(
                  FALSE,
                  TRUE, TRUE, TRUE,
                  FALSE,
                  TRUE, TRUE, TRUE
                )
              ),
              list(showlegend = TRUE)
            )
          )
        )
      )
    )
  )

interactive_plot |> theme_interactive()
```

### Current conclusion

Schools with similar test performance do not have the same predicted HAVO-or-higher advice across levels of school disadvantage. The difference is relatively small at lower test-performance levels and becomes larger as test performance rises. The interactive visualization is the current working version of the final plot; further interpretation, context, and narrative framing can be added later without changing the analysis above.