---
title: "DV Assignment 2: [Your Title Here]"
author: "Group 1: Matheus Nacif, Ilinca Anton, Isabella Braune"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source("../scripts/00-packages.R")

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here("scripts", "01-get-data.R"))
```

## Research question

RQ3. Do schools with similar test results give similar secondary-school advice?

Level: Inference

Target audience: secondary schools, or policymakers.

Instruction: At the end of primary school, every student receives a school advice for a secondary-school track (PRO, VMBO, HAVO, VWO, see the codebook). The advice is based on the teacher’s judgement and on the doorstroomtoets. In this RQ, you will investigate how closely a school’s advice follows its students’ test results, and whether this differs between schools with a more or less disadvantaged student population.

Relate a measure of the school’s advice (e.g. pct_havo_plus) to a measure of its test results (pct_maths_1s and/or pct_reading_2f), and test whether schoolweging moderates this relationship. Do two schools with the same results give the same advice, regardless of who their students are?

### What are we investigating, why does it matter?

Every primary-school pupil receives a secondary-track advice at the end of group 8 (PRO, VMBO, HAVO, or VWO) that sets their entire secondary trajectory. This advice is based on the teacher's judgement and on the doorstroomtoets, a standardized test designed to be comparable across schools. We ask whether, holding test performance constant, a school's advice still depends on how socio-economically disadvantaged its student body is (`schoolweging`). Concretely, we relate each school's share of students advised to HAVO/VWO to its share reaching the reference level in maths and/or reading, and test whether `schoolweging` shifts that relationship. It matters because if disadvantaged schools systematically assign lower tracks to students with the same demonstrated performance, then a standardized test is failing its purpose and students are being pushed down the track on the basis of context rather than ability, a live fairness concern in the Dutch debate over transfer advice.

### Who is the target Audience, how does that shape the choices?

Our primary audience is **policymakers and the general public** (a newspaper-graphic reader), with school boards as a secondary audience. Since this reader is not a statistician, the visualization must carry *one* legible conclusion (advice tracks test results, and disadvantage shifts it by this much) expressed in concrete, relatable units (e.g. percentage points of students advised upward) rather than coefficients or p-values. That pushes us toward a small number of clear encodings (a fit for advantaged vs. disadvantaged schools, or a gap/residual line) and toward a single highlighted pattern, rather than a dense multivariate plot.

#### To Discuss:

We need to check how we want to summarise the variables. Two candidates: (1) `pct_havo_plus` — the share of students advised to HAVO or VWO, a single binary-style metric that is easy for a general audience to read; or (2) a numeric track scale (PRO = 1, VMBO = 2, HAVO = 3, VWO = 6) averaged per school, which uses the full range of the advice and does not collapse the lower tracks (PRO, SBO, VMBO) into one bucket. The choice matters: the binary is simpler to communicate, while the numeric scale is more faithful to how much of a gap a track move represents. We can pick between them once the exploratory plots make clear which measure carries the pattern we want to show.

## Exploratory process

#### Figure 1: Distribution of school weighting across schools

```{r exploratory-plot-1, fig.height = 4, fig.width = 6, fig.align = "center"}
# Drop the 119 empty locations (0 students, no schoolweging)
sw <- schoolweging |>
  filter(!is.na(schoolweging))

sw |>
  ggplot(aes(x = schoolweging)) +
  geom_histogram(bins = 40, fill = "#2878CC", color = "white", alpha = 0.7) +
  geom_density(fill = "#2878CC", alpha = 0.08) +
  geom_vline(
    xintercept = median(sw$schoolweging),
    linetype = "dashed", color = "grey30"
  ) +
  annotate(
    "text",
    x = median(sw$schoolweging) + 0.5,
    y = max(table(cut(sw$schoolweging, breaks = 40))) * 0.85,
    label = paste0("median = ", round(median(sw$schoolweging), 1)),
    hjust = 0, size = 3.5, color = "grey30"
  ) +
  scale_x_continuous(limits = c(18, 43)) +
  labs(
    x = "School weighting (higher = more disadvantaged)",
    y = "Number of schools"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The distribution of `schoolweging` is roughly bell-shaped, centred around 29–30, with the bulk of schools falling between 22 and 37. There is no natural bimodal split — the spread is continuous. For the moderation analysis we'll need to pick a threshold (e.g. median split at \~29.5, or tertiles) to define "low" vs "high" disadvantage groups; the plot makes clear that no single school is an outlier, and that both tails have enough mass to compare.

#### Figure 2: Share of students advised HAVO or higher, by share reaching the reference level

```{r exploratory-plot-2, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}
# the count columns are character ("<5" privacy suppression) -> numeric first
adv_cols <- c("PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT", "VMBO_GT",
              "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO", "ADVIES_NIET_MOGELIJK")
to_num <- function(x) {
  v <- suppressWarnings(as.numeric(x))
  v[is.na(v)] <- 2.5
  v
}

# pull the two id columns out before mutating (their headers carry a
# hidden character and their position differs between the two files)
adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

adv <- schooladviezen |>
  mutate(across(all_of(adv_cols), to_num)) |>
  mutate(
    n_advice = rowSums(across(all_of(adv_cols))),
    pct_havo_plus = 100 * (HAVO + HAVO_VWO + VWO) / n_advice
  )

# proportion of pupils reaching the reference level, averaged across
# maths (1S) and reading (2F) — removes school size as a confound
ref <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_ref = (pct_m + pct_r) / 2
  )


plot_data <- data.frame(inst = adv_ids[[1]], vest = adv_ids[[2]],
                        n_advice = adv$n_advice,
                        pct_havo_plus = adv$pct_havo_plus) |>
  inner_join(
    data.frame(inst = ref_ids[[1]], vest = ref_ids[[2]], pct_ref = ref$pct_ref),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(pct_ref))

plot_data |>
  ggplot(aes(x = pct_ref, y = pct_havo_plus)) +
  geom_point(alpha = 0.3, size = 1, color = "#2878CC") +
  geom_smooth(method = "lm", color = "#EB6734", se = TRUE) +
  labs(
    x = "% of students reaching the reference level (maths 1S + reading 2F)",
    y = "% of students advised to HAVO or higher"
  ) +
  theme_minimal()
```

Figure 2 shows that across the 5,921 schools, the relationship between test result and advice is clearly positive (correlation 0.69): schools whose students reach the reference level in maths and reading advise a larger share of them to HAVO or VWO. On the whole, advice tracks demonstrated performance.

However, our test measure averages two different subjects. Maths and reading measure different skills, and their target levels can differ in difficulty. Averaging them assumes that advice responds to both subjects in the same way.

If advice actually depends more on one subject than the other, averaging them weakens this claim. This would in turn make it unclear what "schools with the same results" means, and that comparison is the core of our research question. We therefore looked at the two subjects separately in figure 3, relating each school's share of VWO advice, the highest track, to its share of students reaching the maths target (1S) and the reading target (2F).

#### Figure 3:

```{r exploratory-plot-3, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}

# share of students reaching the target level, per subject separately
ref_subjects <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) + to_num(REKENEN_1S),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    maths = 100 * to_num(REKENEN_1S) / tot_m,
    reading = 100 * to_num(LV_2F) / tot_r
  )

# one row per school per subject, with readable subject names for the legend
subject_data <- data.frame(
  inst = adv_ids[[1]], vest = adv_ids[[2]],
  n_advice = adv$n_advice,
  pct_vwo = 100 * adv$VWO / adv$n_advice
) |>
  inner_join(
    data.frame(
      inst = ref_ids[[1]], vest = ref_ids[[2]],
      maths = ref_subjects$maths, reading = ref_subjects$reading
    ),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(maths), !is.na(reading)) |>
  pivot_longer(
    c(maths, reading),
    names_to = "subject", values_to = "pct_target"
  ) |>
  mutate(
    label = if_else(subject == "maths", "Maths (1S)", "Reading (2F)")
  ) # Make readable labels for the legend


subject_data |>
  ggplot(
    aes(x = pct_target, y = pct_vwo, colour = label, fill = label)
  ) +
  # shape 21 is a filled circle, so the points use the fill colours
  geom_point(shape = 21, stroke = 0, alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm", formula = y ~ x, se = TRUE,
    alpha = 0.25, linewidth = 1.3
  ) +
  # light colours for the points, darker shades of the same for the lines
  scale_fill_manual(values = c("#009E73", "#CC79A7"), name = NULL) +
  scale_colour_manual(values = c("#00553E", "#7A2E5C"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 25)) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100), expand = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  labs(
    x = "% of students reaching the target level",
    y = "% of students advised VWO"
  ) +
  theme_minimal() +
  theme(legend.position = "top")


# Looking at the correlation:
subject_data |>
  group_by(subject) |>
  summarise(r = cor(pct_target, pct_vwo))
```

Looking at Figure 3, we observe that advice is moderately related to both: slightly more strongly to maths (r = 0.55) than to reading (r = 0.49). Since the difference is small, advice does not appear to depend mainly on one subject. This supports using the average of both subjects as our measure of test results. Because the data is at school level, this describes how schools' advice relates to their overall results, not how individual students are advised.

## Conclusion

School weighting is spread continuously across schools, roughly bell-shaped around 29–30, with enough schools at both ends to compare more and less disadvantaged populations (Figure 1). Schools with better test results clearly advise more students to HAVO or higher (r = 0.69; Figure 2), so on the whole, advice follows demonstrated performance. Looking at the two subjects separately, advice is related to maths (r = 0.55) and reading (r = 0.49) about equally, which supports using the average of both as our measure of test results (Figure 3).

At the same time, figure 2 and 3 both show that schools with similar test results still differ considerably in how many students they advise to a higher track. The question for the next phase is whether schoolweging explains part of that difference. At the same test results, does a more disadvantaged school advise fewer of its students to a higher track?