---
title: "DV Assignment 2: The Hidden Gap Between Test Performance and School Advice"
author: "Group 1: Matheus Nacif, Ilinca Anton, Isabella Braune"
date: today
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source(here::here("scripts", "00-packages.R"))

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here::here("scripts", "01-get-data.R"))

# loads the custom ggplot and plotly themes for this project
# see scripts/02-custom_themes.R for the available functions.
# the theme should only be edited within this file
source(here::here("scripts", "02-custom_themes.R"))
```

## Research question

Do schools with similar test results give similar secondary-school advice?

Level: Inference

Target audience: Policymakers.

Relate a measure of the school’s advice (e.g. pct_havo_plus) to a measure of its test results (pct_maths_1s and/or pct_reading_2f), and test whether schoolweging moderates this relationship. Do two schools with the same results give the same advice, regardless of who their students are?

### Why does it matter weather similar test results give similar secondary-school advice?

Every primary-school pupil receives a secondary-track advice at the end of group 8 (PRO, VMBO, HAVO, or VWO) that sets their entire secondary trajectory. This advice is based on the teacher's judgment and on the doorstroomtoets, a standardized test designed to be comparable across schools. We ask whether, holding test performance constant, a school's advice still depends on how socio-economically disadvantaged its student body is (`schoolweging`). Concretely, we relate each school's share of students advised to HAVO/VWO to its share reaching the reference level in maths and/or reading, and test whether `schoolweging` shifts that relationship. It matters because if disadvantaged schools systematically assign lower tracks to students with the same demonstrated performance, then a standardized test is failing its purpose and students are being pushed down the track on the basis of context rather than ability, a live fairness concern in the Dutch debate over transfer advice.

### Who is the target audience of this report and advice?

Our primary audience is **policymakers and the general public** (a newspaper-graphic reader), with school boards as a secondary audience. Since this reader is not a statistician, the visualization must carry *one* legible conclusion (advice tracks test results, and disadvantage shifts it by this much) expressed in concrete, relatable units (e.g. percentage points of students advised upward) rather than coefficients or p-values. That pushes us toward a small number of clear encodings (a fit for advantaged vs. disadvantaged schools, or a gap/residual line) and toward a single highlighted pattern, rather than a dense multivariate plot.

## Data Processing

Below are the the main variables we're building that will support the rest of the data analysis and visualization for answering the research question. Some things to note about our data processing at this stage:
1. DUO replaces counts below 5 with “\<5” to protect privacy. Dropping these would remove most small schools, so we replace them with 2.5, the midpoint of the possible values 1–4.
2. To calculate the share of students advised HAVO or higher, we divide by all students who received advice. Students for whom no advice was possible, for whatever reason, are excluded (the ADVIES_NIET_MOGELIJK category). Since our outcome is the advice a school gives, these students have no outcome to count. This definition is applied throughout the data processing in this report.


```{r prepare-exploratory-data}

# the count columns are character ("<5" privacy suppression) -> numeric first
advice_cols <- c(
   "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT", "VMBO_GT",
  "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)


to_num <- function(x) {
  if_else(x == "<5", 2.5, suppressWarnings(as.numeric(x)))
}

# Normalize advice headers before preparing either analysis dataset.
names(schooladviezen) <- gsub("\\u00a0", "", names(schooladviezen))

# School identifiers occupy different positions in the two source files.
adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

adv <- schooladviezen |>
  mutate(across(all_of(advice_cols), to_num_exploratory)) |>
  mutate(
    n_advice = rowSums(across(all_of(advice_cols))),
    pct_havo_plus = 100 * (HAVO + HAVO_VWO + VWO) / n_advice
  )

# proportion of pupils reaching the reference level, averaged across
# maths (1S) and reading (2F) — removes school size as a confound
ref <- referentieniveaus |>
  mutate(
    tot_m = to_num_exploratory(REKENEN_LAGER1F) +
      to_num_exploratory(REKENEN_1F) +
      to_num_exploratory(REKENEN_1S) +
      to_num_exploratory(REKENEN_2F),
    tot_r = to_num_exploratory(LV_LAGER1F) +
      to_num_exploratory(LV_1F) + to_num_exploratory(LV_2F),
    pct_m = 100 * to_num_exploratory(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num_exploratory(LV_2F) / tot_r,
    pct_ref = (pct_m + pct_r) / 2
  )

# One school-level dataset for both exploratory advice plots.
exploratory_data <- tibble(
  inst = adv_ids[[1]],
  vest = adv_ids[[2]],
  n_advice = adv$n_advice,
  pct_havo_plus = adv$pct_havo_plus
) |>
  inner_join(
    tibble(
      inst = ref_ids[[1]],
      vest = ref_ids[[2]],
      maths = ref$pct_m,
      reading = ref$pct_r,
      pct_ref = ref$pct_ref
    ),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0)

# Drop the 119 empty locations (0 students, no schoolweging)
sw <- schoolweging |>
  filter(!is.na(schoolweging))

```

### Exploring test results and school advice

Our measure of test results is the percentage of a school’s students who reach the target reference level. Unlike raw test scores, reference levels are comparable across the six test providers. Two subjects have a target level: maths (1S) and reading (2F). Ideally we would summaries them in one measure by averaging the two percentages. That only makes sense if advice responds to both subjects in a similar way, though. The subjects measure different skills, and their target levels may differ in difficulty. If advice depends more on one subject than the other, an average would blur what “schools with the same test results” means, and that comparison is the core of our research question. Figure 1 therefore relates each school’s share of students advised HAVO or higher to its share reaching the maths target and the reading target, separately.

```{r exploratory-plot-subjects, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}

# Dataframe with one row per school per subject
subject_data <- exploratory_data |>
  filter(!is.na(maths), !is.na(reading)) |>
  pivot_longer(
    c(maths, reading), names_to = "subject", values_to = "pct_target"
  ) |>
  mutate(label = if_else(subject == "maths", "Maths (1S)", "Reading (2F)"))

# Compare maths and reading against HAVO+ advice
subject_data |>
  ggplot(
    aes(x = pct_target, y = pct_havo_plus, colour = label, fill = label)
  ) +
  # shape 21 is a filled circle, so the points use the fill colours
  geom_point(shape = 21, stroke = 0, alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm", formula = y ~ x, se = TRUE,
    alpha = 0.25, linewidth = 1.3
  ) +
  # light colours for the points, darker shades of the same for the lines
  scale_fill_manual(values = c("#009E73", "#CC79A7"), name = NULL) +
  scale_colour_manual(values = c("#00553E", "#7A2E5C"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 25)) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100), expand = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  labs(
    title = "Schools whose students reach target levels give more HAVO+ advice",
    subtitle = "Each dot is one school; maths and reading shown separately",
    x = "% of students reaching the target level",
    y = "% of students advised HAVO or higher",
    caption = paste0(
      "Data: DUO, school year 2024–2025. ",
      "Lines: linear fit with 95% interval."
    )
  ) +
  theme_minimal() +
  theme(legend.position = "top")


# correlation and slope per subject on school advice
subject_data |>
  group_by(subject) |>
  summarise(
    r = cor(pct_target, pct_havo_plus),
    slope = coef(lm(pct_havo_plus ~ pct_target))[2]
  )
```

Figure 1 shows that advice relates equally strongly to maths and reading (both r = 0.61). Table 1 shows in a regression with both subjects, their coefficients are practically identical (0.59 each), so advice does not depend more on one subject. The reading line lies lower only because its target is reached more often. We therefore average the two subjects into one test-result measure. Because the data are at school level, this describes how a school’s advice relates to its overall results, not how individual students are advised.

```{r exploratory-plot-average, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}

# Share receiving HAVO+ advice versus average reference-level attainment
plot_data <- exploratory_data |>
  filter(!is.na(pct_ref))

# Plot advice against average test performance
plot_data |>
  ggplot(aes(x = pct_ref, y = pct_havo_plus)) +
  geom_point(alpha = 0.3, size = 1, color = "#2878CC") +
  geom_smooth(method = "lm", color = "#EB6734", se = TRUE) +
  labs(
    title = "Schools with better test results give more HAVO+ advice",
    subtitle = paste0(
      "Each dot is one school. Schools with similar test results ",
      "can still differ widely in their advice."
    ),
    x = paste0(
      "% of students reaching the target level ",
      "(average of maths 1S and reading 2F)"
    ),
    y = "% of students advised HAVO or higher",
    caption = paste0(
      "Data: DUO, school year 2024–2025. Regular primary schools only. ",
      "Line: linear fit with 95% interval."
    )
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(color = "grey50", hjust = 0),
    panel.grid.minor = element_blank()
  )

# linear model: how much HAVO+ advice changes per point of test results
fit_ref <- lm(pct_havo_plus ~ pct_ref, data = plot_data)
summary(fit_ref)

```

Figure 2 shows that across the 5,921 schools, the relationship between test result and advice is clearly positive (correlation 0.69): schools whose students reach the reference level in maths and reading advise a larger share of them to HAVO or VWO. Test results significantly predicted advice, b = \[0.76\], SE = 0.01, t(\[5919\]) = 73.241, p \< .001. This suggests that, a school where 10 percentage points more students reach the target level, ends up advising on average about 7.6 percentage points more of its students to HAVO or higher. Thus, on the whole, advice follows test results. However, test results account for only about half of the differences in advice between schools (R² = 0.48). Schools with the same test results still differed by about 12 percentage points in their advice (residual SE = 11.7). This suggests that factors other than test results also shape the advice a school gives.

### Exploring school weighting

After finding that test results explain only about half of the differences in advice between schools, we turned to school weighting as a possible explanation for the rest. We could add school weighting to the model as a second predictor. However, that would assume that disadvantaged schools differ from other schools by the same amount at every level of test results. Our research question asks whether the relationship between test results and advice itself is the same across schools, regardless of how disadvantaged their students are. We therefore model school weighting as a moderator, by including an interaction between test results and school weighting. This allows the gap in advice between more and less disadvantaged schools to change with test performance, and lets us test whether it does instead of assuming it is constant.

Before testing whether school weighting moderates the relationship between test results and advice, we examined its distribution (Figure 3). We wanted to know whether schools differ enough in school weighting to compare them, and whether some ranges, particularly the most disadvantaged schools, are represented by only a few schools.

```{r exploratory-plot-weighting, fig.height = 4, fig.width = 6, fig.align = "center"}

sw_median <- median(sw$schoolweging)

sw |>
  ggplot(aes(x = schoolweging)) +
  geom_histogram(bins = 40, fill = "#2878CC", color = "white", alpha = 0.7) +
  geom_density(fill = "#2878CC", alpha = 0.08) +
  geom_vline(
    xintercept = sw_median,
    linetype = "dashed", color = "grey30"
  ) +
  annotate(
    "text",
    x = sw_median + 0.5,
    y = max(table(cut(sw$schoolweging, breaks = 40))) * 0.85,
    label = paste0("median = ", round(sw_median, 1)),
    hjust = 0, size = 3.5, color = "grey30"
  ) +
  scale_x_continuous(breaks = seq(20, 40, by = 5)) +
  coord_cartesian(xlim = c(18, 43)) +
  labs(
    title = "Distribution of school weighting",
    subtitle = paste0(
      "Each school gets one score; higher = more disadvantaged ",
      "student population"
    ),
    x = "School weighting",
    y = "Number of schools",
    caption = "Data: Inspectorate of Education / CBS, school year 2024–2025"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The distribution is roughly bell-shaped and centred around 29–30 (median = \[29.5\]), with most schools between 22 and 37. There are no extreme outliers, and both tails contain enough schools to compare less and more disadvantaged schools. The spread is also continuous. There is no natural split into a “low” and a “high” group. Thus for the actual analysis school weighting is a treated as a continoius moderator. However, for the final visualisation we draw lines at the 25thand 75th percentiles of school weighting, representing less and more disadvantaged schools.

## Moderation analysis

For the moderation analysis, we combined the school advice, reference-level attainment, and school weighting datasets. We used the average percentage reaching the target levels in maths (1S) and reading (2F) as our measure of test performance, as established in the exploratory analysis.

### Prepare the analysis data

We refined our measure of HAVO-or-higher advice. The data include recommendations spanning two tracks, such as HAVO/VWO. Rather than assigning the entire HAVO/VWO category to one track, we split its count equally between HAVO and VWO. For our HAVO+ measure, this means counting half of the HAVO/VWO recommendations. This is an approximation, but it allows us to account for mixed advice without treating it as a single-track recommendation.

As before, suppressed counts (`<5`) were replaced with 2.5. We then combined the datasets using school identifiers and excluded schools with missing values for the variables used in the model.

In the exploratory analysis, we only looked at the distribution of school weighting on its own, without linking it to the other data. For the model, we link it to the advice and test data. The three DUO files share the school code (INSTELLINGSCODE) and location code (VESTIGINGSCODE), but the school weighting file uses its own identifier (OVT), formatted like “00AP|C1”. The part before the “|” is the school code,and the part after does not follow the same location numbering as the DUO files.We therefore match on the school code only. When a school has several locations with a school weighting, we keep the weighting of the first location listed and assign it to all locations of that school. This affects 305 out of 5876 schools. For most of them, the difference is small: locations of the same school differ by a median of 2.2 points in school weighting, compared to a range of roughly 22–37 across schools. A few schools, however, have locations that differ by up to 14.7 points, so for those schools the assigned weighting may not reflect every location well. Finally, schools without a disadvantage weighting are exluded from the data. 

```{r prepare-moderation-data}
advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)

# Only suppressed values are imputed here; other missing values remain NA.
to_num <- function(x) {
  if_else(x == "<5", 2.5, suppressWarnings(as.numeric(x)))
}

# Getting the advice data
advice_data <- schooladviezen |>
  mutate(
    inst = adv_ids[[1]],
    vest = adv_ids[[2]],
    across(all_of(advice_cols), to_num),
    n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice
  ) |>
  filter(n_advice > 0) |>
  select(inst, vest, pct_havo_plus)

reference_data <- referentieniveaus |>
  mutate(
    inst = ref_ids[[1]],
    vest = ref_ids[[2]],
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  ) |>
  select(inst, vest, pct_test)

# Schoolweighting data
weighting_data <- schoolweging |>
  transmute(
    inst = sub("\\|.*$", "", OVT),
    schoolweging
  ) |>
  filter(!is.na(schoolweging)) |>
  distinct(inst, .keep_all = TRUE)

# how many schools in the model data are affected by the data processing choice
multi_location <- schoolweging |>
  filter(!is.na(schoolweging)) |>
  count(inst = sub("\\|.*$", "", OVT)) |>
  filter( n > 1)
# sum(model_data$inst %in% multi_location$inst)

# how much do locations of the same school differ in school weighting?
schoolweging |>
  filter(!is.na(schoolweging)) |>
  mutate(inst = sub("\\|.*$", "", OVT)) |>
  group_by(inst) |>
  filter(n() > 1) |>
  summarise(spread = max(schoolweging) - min(schoolweging)) |>
  summarise(median_spread = median(spread), max_spread = max(spread))

# The final model data
model_data <- advice_data |>
  inner_join(reference_data, by = c("inst", "vest")) |>
  inner_join(weighting_data, by = "inst") |>
  filter(!is.na(pct_test), !is.na(pct_havo_plus))

cat("Number of rows in the moderation dataframe:", nrow(model_data))
head(model_data)
```

```{r}

# how much do locations of the same school differ in school weighting?
schoolweging |>
  filter(!is.na(schoolweging)) |>
  mutate(inst = sub("\\|.*$", "", OVT)) |>
  group_by(inst) |>
  filter(n() > 1) |>
  summarise(spread = max(schoolweging) - min(schoolweging)) |>
  summarise(median_spread = median(spread), max_spread = max(spread))
```


### Model school disadvantage as a moderator

We fitted a linear regression with an interaction between test performance and school weighting. As discussed earlier, this allows us to examine whether the relationship between test results and advice changes depending on how disadvantaged a school's student population is.

```{r moderation-model}
model_havo <- lm(pct_havo_plus ~ pct_test * schoolweging, data = model_data)

model_results <- tibble(
  r_squared = summary(model_havo)$r.squared,
  interaction = coef(model_havo)[["pct_test:schoolweging"]]
)

model_results
```

The model explained approximately 53% of the variation in HAVO-or-higher advice between schools (R² = 0.53). The interaction coefficient was negative (b = −0.033), suggesting that the positive relationship between test performance and HAVO+ advice becomes weaker as school disadvantage increases.

In other words, schools with lower disadvantage show a steeper increase in HAVO+ advice as test performance rises. For more disadvantaged schools, this increase is smaller. This means that the predicted difference in advice between the two becomes larger at higher levels of test performance.

#### Representative predictions

Although the interaction tells us how the relationship changes, its coefficient is not particularly intuitive. We therefore calculated model predictions at two representative levels of school disadvantage: the 25th and 75th percentiles of `schoolweging` (approximately 27.4 and 31.8).

We chose these values because they represent meaningful differences in disadvantage without relying on the extremes of the distribution. Importantly, these are only reference values for the visualization. School weighting remains continuous in the model.

```{r moderation-predictions}
sw_values <- quantile(
  model_data$schoolweging,
  probs = c(0.25, 0.50, 0.75),
  na.rm = TRUE
)

```

To see what this means in practice, we compared the predicted percentage of HAVO+ advice at different levels of test performance.

```{r moderation-gap-check}
model_data <- model_data |>
  mutate(
    disadvantage_group = case_when(
      schoolweging <= sw_values[[1]] ~ "Lower disadvantage",
      schoolweging >= sw_values[[3]] ~ "Higher disadvantage",
      TRUE ~ NA_character_
    ),
    disadvantage_group = factor(
      disadvantage_group,
      levels = c("Lower disadvantage", "Higher disadvantage")
    )
  )

# Extend the fitted lines without changing the observed data.
x_line <- seq(10, 90, length.out = 200)

prediction_interactive <- expand_grid(
  pct_test = x_line,
  disadvantage = c("Lower disadvantage", "Higher disadvantage")
) |>
  mutate(
    schoolweging = if_else(
      disadvantage == "Lower disadvantage",
      as.numeric(sw_values[[1]]),
      as.numeric(sw_values[[3]])
    ),
    fit = predict(model_havo, newdata = pick(everything()))
  )

prediction_interactive <- prediction_interactive |>
  group_by(pct_test) |>
  mutate(
    difference_high = fit - fit[disadvantage == "Higher disadvantage"]
  ) |>
  ungroup() |>
  mutate(
    hover_text = paste0(
      "<b>", disadvantage, "</b>",
      "<br><br>Test performance: ", round(pct_test, 1), "%",
      "<br>Predicted HAVO+ advice: ", round(fit, 1), "%",
      "<br><br><b>Compared with higher-disadvantage schools:</b>",
      "<br><b>",
      if_else(
        abs(difference_high) < 1e-8,
        "Reference group",
        paste0(
          if_else(difference_high > 0, "+", ""),
          round(difference_high, 1),
          " percentage points"
        )
      ),
      "</b>"
    )
  )

gap_70 <- predict(
  model_havo,
  newdata = tibble(
    pct_test = 70,
    schoolweging = as.numeric(sw_values[c(1, 3)])
  )
) |>
  diff() |>
  abs() |>
  round(1)

overall_model <- lm(pct_havo_plus ~ pct_test, data = model_data)

overall_data <- tibble(pct_test = x_line) |>
  mutate(fit = predict(overall_model, newdata = pick(everything())))

overall_model
overall_data
```

The predicted gap is relatively small when test performance is low but becomes wider as performance increases. For example, when 70% of pupils reach the target reference levels, the model predicts approximately 47% HAVO+ advice at the lower-disadvantage value, compared with 39% at the higher-disadvantage value. That is a difference of around 8 percentage points, despite the same overall test performance.

This is the main finding we wanted to communicate in our visualization: similar test results do not necessarily correspond to similar advice, and the differences appear to increase at higher levels of performance.

### Interactive visualization

For the final visualization, we wanted to make this difference visible without requiring readers to interpret regression coefficients.

We decided to start with the overall relationship between test performance and HAVO+ advice as it is somewhat easy to interpret. The interactive button then introduces school disadvantage, replacing the overall regression line with two predictions from the moderation model. Our intention was that the reader/viewer could uncover the differences based on school disadvantage themselves.

We chose to show only the 25th and 75th percentiles rather than several lines. This makes the widening gap easier to see without overcrowding the graph. The individual schools remain visible to show the variation around the predictions, with the lower- and higher-disadvantage groups highlighted when the comparison is activated.

We also added hover information showing the predicted HAVO+ percentage and its difference from the higher-disadvantage reference. This allows readers to explore the size of the gap in percentage points rather than estimating it visually. During testing, the base hover option on the plot was quite unreliable as it requires the reader to point exactly at the line and it does not work in some angles. Terefore, we recommend using the secondary option that picks up the closest hover information.

```{r interactive-data}
# Title and subtitles.
title_text <- paste0(
  "<b>The <span style='color:", group_colors[["Higher disadvantage"]],
  "'>Hidden Gap</span> Between Test Performance",
  "<br>and School Advice</b>"
)

subtitle_overall <- paste0(
  "Schools with better test results tend to give more HAVO+ advice.",
  "<br>But does this relationship hold equally across schools?"
)

subtitle_gap <- paste0(
  "At schools where 70% of pupils reach the target level,",
  "<br><b>about ", round(gap_70), " fewer pupils per 100</b> ",
  "are predicted to receive HAVO+ advice",
  "<br>in <span style='color:", group_colors[["Higher disadvantage"]],
  "'><b>higher-disadvantage schools</b></span> than in ",
  "<span style='color:", group_colors[["Lower disadvantage"]],
  "'><b>lower-disadvantage schools</b></span>."
)

# Methodological notes and attribution.
footnote_1 <- paste0(
  "<b>Additional information:</b> Each dot represents one school (N = ",
  format(nrow(model_data), big.mark = ","),
  "). Test results combine the share of pupils reaching target levels",
  "<br>in maths and reading. Lower and higher disadvantage refer to the bottom",
  " and top 25% of schools by schoolweging. The lines are",
  "<br>model predictions at the 25th and 75th percentiles.",
  " Axes display 0-100% for test attainment and 0-80% for HAVO-or-higher ",
  "advice."
)

footnote_2 <- paste0(
  "<b>Source:</b> DUO, school year 2024-2025; ",
  "schoolweging (2022-2025).",
  "<br><b>Authors:</b> Matheus Nacif, Ilinca Anton, Isabella Braune."
)

interactive_plot <- plot_ly(height = 850)

# Default state: all schools
interactive_plot <- interactive_plot |>
  add_markers(
    data = model_data,
    x = ~pct_test,
    y = ~pct_havo_plus,
    marker = list(color = group_neutral_col, size = 5, opacity = 0.35),
    hoverinfo = "skip",
    showlegend = FALSE
  )

# Lower- and higher-disadvantage dots
for (group in c("Lower disadvantage", "Higher disadvantage")) {
  group_data <- model_data |>
    filter(disadvantage_group == group)

  interactive_plot <- interactive_plot |>
    add_markers(
      data = group_data,
      x = ~pct_test,
      y = ~pct_havo_plus,
      name = group,
      marker = list(
        color = group_colors[[group]],
        size = 5,
        opacity = 0.22
      ),
      hoverinfo = "skip",
      visible = FALSE,
      showlegend = FALSE,
      legendgroup = group
    )
}

# Overall fitted relationship
interactive_plot <- interactive_plot |>
  add_lines(
    data = overall_data,
    x = ~pct_test,
    y = ~fit,
    line = list(color = group_sec_text, width = 5),
    hovertemplate = paste0(
      "<b>Overall relationship</b>",
      "<br><br>Test performance: %{x:.1f}%",
      "<br>Predicted HAVO+ advice: %{y:.1f}%",
      "<extra></extra>"
    ),
    showlegend = FALSE
  )

# Lower- and higher-disadvantage predictions
for (group in c("Lower disadvantage", "Higher disadvantage")) {
  line_data <- prediction_interactive |>
    filter(disadvantage == group)

  interactive_plot <- interactive_plot |>
    add_lines(
      data = line_data,
      x = ~pct_test,
      y = ~fit,
      name = group,
      text = ~hover_text,
      hovertemplate = "%{text}<extra></extra>",
      line = list(color = group_colors[[group]], width = 5),
      visible = FALSE,
      showlegend = TRUE,
      legendgroup = group
    )
}

interactive_plot <- interactive_plot |>
  layout(
    margin = list(l = 85, r = 25, t = 225, b = 240),

    annotations = list(
      # Title
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "bottom",
        align = "left",
        yshift = 112,
        text = title_text,
        showarrow = FALSE,
        font = list(size = group_px$title, color = group_prim_text)
      ),

      # Dynamic subtitle (annotation index 1)
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "bottom",
        align = "left",
        yshift = 58,
        text = subtitle_overall,
        showarrow = FALSE,
        font = list(size = group_px$subtitle, color = group_sec_text)
      ),

      # Hover instruction
      list(
        x = 1,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "right",
        yanchor = "bottom",
        yshift = 8,
        text = "Hover over the lines to compare predictions",
        showarrow = FALSE,
        font = list(size = 12, color = group_sec_text)
      ),

      # Additional information
      list(
        x = 0,
        y = 0,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        yshift = -105,
        align = "left",
        text = footnote_1,
        showarrow = FALSE,
        font = list(size = 11, color = group_sec_text)
      ),

      # Source and authors
      list(
        x = 0,
        y = 0,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        yshift = -160,
        align = "left",
        text = footnote_2,
        showarrow = FALSE,
        font = list(size = 11, color = group_sec_text)
      )
    ),

    xaxis = list(
      title = "Pupils reaching the target test level (%)",
      range = c(0, 100),
      tickvals = seq(0, 100, 20),
      showline = TRUE,
      automargin = TRUE
    ),

    yaxis = list(
      title = "Pupils receiving HAVO or higher advice (%)",
      range = c(0, 80),
      tickvals = seq(0, 80, 20),
      showline = TRUE,
      automargin = TRUE
    ),

    legend = list(
      orientation = "h",
      x = 0.5,
      y = 0.14,
      xanchor = "center",
      yanchor = "middle"
    ),

    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0,
        y = 1,
        xanchor = "left",
        yanchor = "bottom",
        showactive = TRUE,
        buttonwidth = 190,
        bgcolor = "#EAE1D4",
        bordercolor = "#B9AA94",
        borderwidth = 1,
        font = list(size = 14, color = group_prim_text),
        pad = list(b = 12),

        buttons = list(
          list(
            label = "Overall relationship",
            method = "update",
            args = list(
              list(
                visible = c(TRUE, FALSE, FALSE, TRUE, FALSE, FALSE)
              ),
              list(
                showlegend = FALSE,
                "annotations[1].text" = subtitle_overall
              )
            )
          ),

          list(
            label = "By disadvantage",
            method = "update",
            args = list(
              list(
                visible = c(FALSE, TRUE, TRUE, FALSE, TRUE, TRUE)
              ),
              list(
                showlegend = TRUE,
                "annotations[1].text" = subtitle_gap
              )
            )
          )
        )
      )
    )
  )

interactive_plot |>
  theme_interactive()

```

### Conclusion

Our results suggest that school disadvantage is associated with how closely secondary-school advice follows test performance. Although schools with better test results generally give more HAVO+ advice, this relationship is weaker in more disadvantaged schools. The difference is particularly noticeable at higher levels of test performance.

However, these are school-level comparisons. We cannot conclude that individual pupils with identical test results receive different advice because of their socioeconomic background. Other differences between schools may also contribute to the pattern, especially considering the differences in test providers and how these are interpreted, which is an important debate now.

Nevertheless, the analysis shows that test performance alone does not fully explain differences in secondary-school advice, and that school disadvantage is relevant to understanding them.