---
title: "DV Assignment 2: [Your Title Here]"
author: "Group 1: Matheus Nacif, Ilinca Anton, Isabella Braune"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clear environment so knitting only depends on what's in this file
rm(list = ls())

# installs (if missing) and loads every package this project uses - add any
# new packages to scripts/00-packages.R, not here
source(here::here("scripts", "00-packages.R"))

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging into this
# session - see scripts/01-get-data.R for what each one contains
source(here::here("scripts", "01-get-data.R"))

# loads the custom ggplot and plotly themes for this project 
# see scripts/02-custom_themes.R for the available functions. 
# the theme should only be edited within this file 
source(here::here("scripts", "02-custom_themes.R"))
```

## Research question

Do schools with similar test results give similar secondary-school advice?

Level: Inference

Target audience: Policymakers.

Relate a measure of the school’s advice (e.g. pct_havo_plus) to a measure of its test results (pct_maths_1s and/or pct_reading_2f), and test whether schoolweging moderates this relationship. Do two schools with the same results give the same advice, regardless of who their students are?

### Why does it matter weather similar test results give similar secondary-school advice?

Every primary-school pupil receives a secondary-track advice at the end of group 8 (PRO, VMBO, HAVO, or VWO) that sets their entire secondary trajectory. This advice is based on the teacher's judgement and on the doorstroomtoets, a standardized test designed to be comparable across schools. We ask whether, holding test performance constant, a school's advice still depends on how socio-economically disadvantaged its student body is (`schoolweging`). Concretely, we relate each school's share of students advised to HAVO/VWO to its share reaching the reference level in maths and/or reading, and test whether `schoolweging` shifts that relationship. It matters because if disadvantaged schools systematically assign lower tracks to students with the same demonstrated performance, then a standardized test is failing its purpose and students are being pushed down the track on the basis of context rather than ability, a live fairness concern in the Dutch debate over transfer advice.

### Who is the target audience of this report and advcie?

Our primary audience is **policymakers and the general public** (a newspaper-graphic reader), with school boards as a secondary audience. Since this reader is not a statistician, the visualization must carry *one* legible conclusion (advice tracks test results, and disadvantage shifts it by this much) expressed in concrete, relatable units (e.g. percentage points of students advised upward) rather than coefficients or p-values. That pushes us toward a small number of clear encodings (a fit for advantaged vs. disadvantaged schools, or a gap/residual line) and toward a single highlighted pattern, rather than a dense multivariate plot.

## Data Processing
Below are the the main variables we're building that will support the rest of the data analysis and visualization for answering the research question.

```{r, data processing}

# the count columns are character ("<5" privacy suppression) -> numeric first
adv_cols <- c("PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT", "VMBO_GT",
              "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO", "ADVIES_NIET_MOGELIJK")

to_num <- function(x) {
  v <- suppressWarnings(as.numeric(x))
  v[is.na(v)] <- 2.5
  v
}

# pull the two id columns out before mutating (their headers carry a
# hidden character and their position differs between the two files)
adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

adv <- schooladviezen |>
  mutate(across(all_of(adv_cols), to_num)) |>
  mutate(
    n_advice = rowSums(across(all_of(adv_cols))),
    pct_havo_plus = 100 * (HAVO + HAVO_VWO + VWO) / n_advice
  )

# proportion of pupils reaching the reference level, averaged across
# maths (1S) and reading (2F) — removes school size as a confound
ref <- referentieniveaus |>
  mutate(
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_ref = (pct_m + pct_r) / 2    # the average, will be used later
  )

# Drop the 119 empty locations (0 students, no schoolweging)
sw <- schoolweging |>
  filter(!is.na(schoolweging))

```


### Looking at the outcome variable: Secondary-school Advice

### Looking at the predictor: Test Results

Our measure of test results is the percentage of a school’s students who reach the target reference level. Unlike raw test scores, reference levels are comparable across the six test providers. Two subjects have a target level: maths (1S) and reading (2F). Ideally we would summaries them in one measure by averaging the two percentages. That only makes sense if advice responds to both subjects in a similar way, though. The subjects measure different skills, and their target levels may differ in difficulty. If advice depends more on one subject than the other, an average would blur what “schools with the same test results” means, and that comparison is the core of our research question. Figure 1 therefore relates each school’s share of students advised HAVO or higher to its share reaching the maths target and the reading target, separately.
 
```{r, name something}

# one row per school per subject, with readable subject names for the legend
# wide version: one row per school, both subjects as columns
subject_wide <- data.frame(
  inst = adv_ids[[1]], vest = adv_ids[[2]],
  n_advice = adv$n_advice,
  pct_havo_plus = adv$pct_havo_plus
) |>
  inner_join(
    data.frame(
      inst = ref_ids[[1]], vest = ref_ids[[2]],
      maths = ref$pct_m, reading = ref$pct_r
    ),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(maths), !is.na(reading))

#long version: one row per school per subject, with readable subject names for the legend
subject_data <- subject_wide |>
  pivot_longer(c(maths, reading), names_to = "subject", values_to = "pct_target") |>
  mutate(label = if_else(subject == "maths", "Maths (1S)", "Reading (2F)"))

# Plotting reading and math target levels on to school advice
subject_data |>
  ggplot(
    aes(x = pct_target, y = pct_havo_plus, colour = label, fill = label)
  ) +
  # shape 21 is a filled circle, so the points use the fill colours
  geom_point(shape = 21, stroke = 0, alpha = 0.3, size = 1.3) +
  geom_smooth(
    method = "lm", formula = y ~ x, se = TRUE,
    alpha = 0.25, linewidth = 1.3
  ) +
  # light colours for the points, darker shades of the same for the lines
  scale_fill_manual(values = c("#009E73", "#CC79A7"), name = NULL) +
  scale_colour_manual(values = c("#00553E", "#7A2E5C"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 25)) +
  scale_y_continuous(breaks = seq(0, 100, by = 25)) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100), expand = FALSE) +
  guides(fill = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  labs(
  title = "Schools whose students reach target levels give more HAVO+ advice",
  subtitle = "Each dot is one school; maths and reading shown separately",
  x = "% of students reaching the target level",
  y = "% of students advised HAVO or higher",
  caption = "Data: DUO, school year 2024–2025. Lines: linear fit with 95% interval."
) +
  theme_minimal() +
  theme(legend.position = "top")


# correlation and slope per subject on school advice
subject_data |>
  group_by(subject) |>
  summarise(
    r     = cor(pct_target, pct_havo_plus),
    slope = coef(lm(pct_havo_plus ~ pct_target))[2]
  ) 
```

Figure 1 shows that advice relates equally strongly to maths and reading (both r = 0.61). Table 1 shows in a regression with both subjects, their coefficients are practically identical (0.38 each), so advice does not depend more on one subject. The reading line lies lower only because its target is reached more often. We therefore average the two subjects into one test-result measure. Because the data are at school level, this describes how a school’s advice relates to its overall results, not how individual students are advised.


```{r exploratory-plot-2, fig.height = 4.5, fig.width = 6.5, fig.align = "center"}

# Data for plotting the share of students advised HAVO or higher, by share reaching the reference level
plot_data <- data.frame(inst = adv_ids[[1]], 
                        vest = adv_ids[[2]],
                        n_advice = adv$n_advice,
                        pct_havo_plus = adv$pct_havo_plus) |>
  inner_join(
    data.frame(inst = ref_ids[[1]], vest = ref_ids[[2]], pct_ref = ref$pct_ref),
    by = c("inst", "vest")
  ) |>
  filter(n_advice > 0, !is.na(pct_ref))

# Plotting the share of students advised HAVO or higher, by share reaching the reference level
plot_data |>
  ggplot(aes(x = pct_ref, y = pct_havo_plus)) +
  geom_point(alpha = 0.3, size = 1, color = "#2878CC") +
  geom_smooth(method = "lm", color = "#EB6734", se = TRUE) +
  labs(
    title = "Schools with better test results give more HAVO+ advice",
    subtitle = "Each dot is one school. Schools with similar test results can still differ widely in their advice.",
    x = "% of students reaching the target level (average of maths 1S and reading 2F)",
    y = "% of students advised HAVO or higher",
    caption = "Data: DUO, school year 2024–2025. Regular primary schools only. Line: linear fit with 95% interval."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(color = "grey50", hjust = 0),
    panel.grid.minor = element_blank()
  )

# linear model: how much HAVO+ advice changes per point of test results
fit_ref <- lm(pct_havo_plus ~ pct_ref, data = plot_data)
summary(fit_ref)

```

Figure 2 shows that across the 5,921 schools, the relationship between test result and advice is clearly positive (correlation 0.69): schools whose students reach the reference level in maths and reading advise a larger share of them to HAVO or VWO. Test results significantly predicted advice, b = [0.76], SE = 0.01, t([5919]) = 73.241, p < .001. This suggests that, a school where 10 percentage points more students reach the target level, ends up advising on average about 7.6 percentage points more of its students to HAVO or higher. Thus, on the whole, advice follows test results. However, test results account for only about half of the differences in advice between schools (R² = 0.48). Schools with the same test results still differed by about 12 percentage points in their advice (residual SE = 11.7). This suggests that factors other than test results also shape the advice a school gives. The next step is to investigate weather, a school's school weighting score could be such a factor. 

### Looking at the moderator: School Weighting
After finding that test results explain only about half of the differences in advice between schools, we turned to school weighting as a possible explanation for the rest. Before testing whether school weighting moderates the relationship between test results and advice, we examined its distribution (Figure 3). We wanted to know whether schools differ enough in school weighting to compare them, and whether some ranges, particularly the most disadvantaged schools, are represented by only a few schools.

```{r exploratory-plot-1, fig.height = 4, fig.width = 6, fig.align = "center"}

sw |>
  ggplot(aes(x = schoolweging)) +
  geom_histogram(bins = 40, fill = "#2878CC", color = "white", alpha = 0.7) +
  geom_density(fill = "#2878CC", alpha = 0.08) +
  geom_vline(
    xintercept = median(sw$schoolweging),
    linetype = "dashed", color = "grey30"
  ) +
  annotate(
    "text",
    x = median(sw$schoolweging) + 0.5,
    y = max(table(cut(sw$schoolweging, breaks = 40))) * 0.85,
    label = paste0("median = ", round(median(sw$schoolweging), 1)),
    hjust = 0, size = 3.5, color = "grey30"
  ) +
  scale_x_continuous(breaks = seq(20, 40, by = 5)) +
  coord_cartesian(xlim = c(18, 43)) +
  labs(
    title = "Distribution of school weighting",
    subtitle = "Each school gets one score; higher = more disadvantaged student population",
    x = "School weighting",
    y = "Number of schools",
    caption = "Data: Inspectorate of Education / CBS, school year 2024–2025"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The distribution is roughly bell-shaped and centred around 29–30 (median = [29.5]), with most schools between 22 and 37. There are no extreme outliers, and both tails contain enough schools to compare less and more disadvantaged schools. The spread is also continuous: there is no natural split into a “low” and a “high” group. To show the moderation visually, we therefore have to choose a cut-off ourselves, such as a median split at about 29.5 or three equal-sized groups (tertiles).


## Moderation analysis

The exploratory analysis above supports using the average of maths 1S and reading 2F as the measure of school-level test performance. We now test whether `schoolweging` changes the relationship between test performance and HAVO-or-higher advice.

For this analysis:

- privacy-suppressed counts (`"<5"`) are replaced by **2.5**;
- split-track advice (`HAVO_VWO`) contributes **0.5** to the HAVO-or-higher outcome;
- `schoolweging` remains continuous in the model;
- the 25th, 50th, and 75th percentiles of `schoolweging` are used only to visualize representative levels of disadvantage.

### Prepare the analysis data

```{r prepare-moderation-data}
names(schooladviezen) <- gsub("\\u00a0", "", names(schooladviezen))

advice_cols <- c(
  "VSO", "PRO", "VMBO_B", "VMBO_B_K", "VMBO_K", "VMBO_K_GT",
  "VMBO_GT", "VMBO_GT_HAVO", "HAVO", "HAVO_VWO", "VWO"
)

to_num <- function(x) {
  if_else(x == "<5", 2.5, suppressWarnings(as.numeric(x)))
}

adv_ids <- schooladviezen[, 3:4]
ref_ids <- referentieniveaus[, 2:3]

advice_data <- schooladviezen |>
  mutate(
    inst = adv_ids[[1]],
    vest = adv_ids[[2]],
    across(all_of(advice_cols), to_num),
    n_advice = rowSums(across(all_of(advice_cols)), na.rm = TRUE),
    pct_havo_plus = 100 * (HAVO + 0.5 * HAVO_VWO + VWO) / n_advice
  ) |>
  filter(n_advice > 0) |>
  select(inst, vest, pct_havo_plus)

reference_data <- referentieniveaus |>
  mutate(
    inst = ref_ids[[1]],
    vest = ref_ids[[2]],
    tot_m = to_num(REKENEN_LAGER1F) + to_num(REKENEN_1F) +
      to_num(REKENEN_1S) + to_num(REKENEN_2F),
    tot_r = to_num(LV_LAGER1F) + to_num(LV_1F) + to_num(LV_2F),
    pct_m = 100 * to_num(REKENEN_1S) / tot_m,
    pct_r = 100 * to_num(LV_2F) / tot_r,
    pct_test = (pct_m + pct_r) / 2
  ) |>
  select(inst, vest, pct_test)

weighting_data <- schoolweging |>
  transmute(
    inst = sub("\\|.*$", "", OVT),
    schoolweging
  ) |>
  filter(!is.na(schoolweging)) |>
  distinct(inst, .keep_all = TRUE)

model_data <- advice_data |>
  inner_join(reference_data, by = c("inst", "vest")) |>
  inner_join(weighting_data, by = "inst") |>
  filter(!is.na(pct_test), !is.na(pct_havo_plus))

nrow(model_data)
```

The analysis is performed at the **school level**. The outcome is the percentage receiving HAVO-or-higher advice, while test performance is based on standardized reference-level attainment rather than provider-specific raw scores.

### Model school disadvantage as a moderator

The interaction model tests whether the relationship between test performance and HAVO-or-higher advice changes with school disadvantage.

```{r moderation-model}
model_havo <- lm(
  pct_havo_plus ~ pct_test * schoolweging,
  data = model_data
)

model_results <- tibble(
  r_squared = summary(model_havo)$r.squared,
  interaction = coef(model_havo)[["pct_test:schoolweging"]]
)

model_results
```

The interaction is negative, meaning that the positive relationship between test performance and HAVO-or-higher advice becomes weaker as school disadvantage increases. In other words, the difference between lower- and higher-disadvantage schools becomes larger at higher levels of test performance.

#### Representative predictions

The model keeps `schoolweging` continuous. The three lines used in the visualization represent its 25th, 50th, and 75th percentiles.

```{r moderation-predictions}
sw_values <- quantile(
  model_data$schoolweging,
  probs = c(0.25, 0.50, 0.75),
  na.rm = TRUE
)

prediction_data <- expand_grid(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  ),
  schoolweging = as.numeric(sw_values)
) |>
  mutate(
    disadvantage = factor(
      schoolweging,
      levels = as.numeric(sw_values),
      labels = c("Lower disadvantage", "Typical", "Higher disadvantage")
    ),
    fit = predict(model_havo, newdata = pick(pct_test, schoolweging))
  )
```

To make the widening gap easier to interpret, we compare the lower- and higher-disadvantage predictions at several test-performance values.

```{r moderation-gap-check}
model_data <- model_data |>
  mutate(
    disadvantage_group = case_when(
      schoolweging <= sw_values[[1]] ~ "Lower disadvantage",
      schoolweging >= sw_values[[3]] ~ "Higher disadvantage",
      TRUE ~ NA_character_
    ),
    disadvantage_group = factor(
      disadvantage_group,
      levels = c("Lower disadvantage", "Higher disadvantage")
    )
  )

high_reference <- prediction_data |>
  filter(disadvantage == "Higher disadvantage") |>
  select(pct_test, high_fit = fit)

prediction_interactive <- prediction_data |>
  filter(disadvantage %in% c("Lower disadvantage", "Higher disadvantage")) |>
  left_join(high_reference, by = "pct_test") |>
  mutate(
    difference_high = fit - high_fit,
    hover_text = paste0(
      "<b>", disadvantage, "</b>",
      "<br><br>Test performance: ", round(pct_test, 1), "%",
      "<br>Predicted HAVO+ advice: ", round(fit, 1), "%",
      "<br><br><b>Compared with higher-disadvantage schools:</b>",
      "<br><b>",
      if_else(
        abs(difference_high) < 1e-8,
        "Reference group",
        paste0(
          if_else(difference_high > 0, "+", ""),
          round(difference_high, 1),
          " percentage points"
        )
      ),
      "</b>"
    )
  )


gap_70 <- predict(
  model_havo,
  newdata = tibble(
    pct_test = 70,
    schoolweging = as.numeric(sw_values[c(1, 3)])
  )
) |>
  diff() |>
  abs() |>
  round(1)


overall_model <- lm(pct_havo_plus ~ pct_test, data = model_data)

overall_data <- tibble(
  pct_test = seq(
    quantile(model_data$pct_test, 0.05, na.rm = TRUE),
    quantile(model_data$pct_test, 0.95, na.rm = TRUE),
    length.out = 100
  )
) |>
  mutate(fit = predict(overall_model, newdata = pick(everything())))
```

<<<<<<< HEAD
```{r interactive-plot}
# removed the call for group_colours here 
=======
The predicted difference is small at lower test-performance levels and widens as test performance rises. At 70% test performance, the predicted HAVO-or-higher advice gap between the 25th and 75th percentiles of school disadvantage is about **8 percentage points**.
>>>>>>> d0ca5faa28835aa739bfcd8bda77ac3cdbefaab4

### Interactive visualization

The interactive plot has two states:

1. **Overall:** observed schools and the overall relationship between test performance and HAVO-or-higher advice.
2. **Show school disadvantage:** schools are coloured by broad disadvantage group and the overall line is replaced by predictions at the 25th, 50th, and 75th percentiles of `schoolweging`.

The broad groups used to colour the points are only a visual aid. The model itself continues to use continuous `schoolweging`.

```{r interactive-data}
# Title and subtitles
title_text <- paste0(
  "<b>The <span style='color:", group_colors[["Higher disadvantage"]],
  "'>Hidden Gap</span> Between Test Performance",
  "<br>and School Advice</b>"
)

subtitle_overall <- paste0(
  "Schools with better test results tend to give more HAVO+ advice.",
  "<br>But does this relationship hold equally across schools?"
)

subtitle_gap <- paste0(
  "At schools where 70% of pupils reach the target level,",
  "<br><b>about ", round(gap_70), " fewer pupils per 100</b> ",
  "are predicted to receive HAVO+ advice",
  "<br>in <span style='color:", group_colors[["Higher disadvantage"]],
  "'><b>higher-disadvantage schools</b></span> than in ",
  "<span style='color:", group_colors[["Lower disadvantage"]],
  "'><b>lower-disadvantage schools</b></span>."
)

# Methodological notes and attribution
footnote_1 <- paste0(
  "<b>Additional information:</b> Each dot represents one school (N = ",
  format(nrow(model_data), big.mark = ","),
  "). Test results combine the share of pupils reaching target levels",
  "<br>in maths and reading. Lower and higher disadvantage refer to the bottom",
  " and top 25% of schools by schoolweging. The lines are model predictions",
  "<br>at the 25th and 75th percentiles. Axes display 0-100% for test",
  " attainment and 0-80% for HAVO-or-higher advice."
)

footnote_2 <- paste0(
  "<b>Source:</b> DUO, school year 2024-2025; ",
  "schoolweging (2022-2025).",
  "<br><b>Authors:</b> Matheus Nacif, Ilinca Anton, Isabella Braune."
)

interactive_plot <- plot_ly(height = 850)

# Default state: all schools
interactive_plot <- interactive_plot |>
  add_markers(
    data = model_data,
    x = ~pct_test,
    y = ~pct_havo_plus,
    marker = list(color = group_neutral_col, size = 5, opacity = 0.35),
    hoverinfo = "skip",
    showlegend = FALSE
  )

# Lower- and higher-disadvantage dots
for (group in c("Lower disadvantage", "Higher disadvantage")) {
  group_data <- model_data |>
    filter(disadvantage_group == group)

  interactive_plot <- interactive_plot |>
    add_markers(
      data = group_data,
      x = ~pct_test,
      y = ~pct_havo_plus,
      name = group,
      marker = list(
        color = group_colors[[group]],
        size = 5,
        opacity = 0.22
      ),
      hoverinfo = "skip",
      visible = FALSE,
      showlegend = FALSE,
      legendgroup = group
    )
}

# Overall fitted relationship
interactive_plot <- interactive_plot |>
  add_lines(
    data = overall_data,
    x = ~pct_test,
    y = ~fit,
    line = list(color = group_sec_text, width = 5),
    hovertemplate = paste0(
      "<b>Overall relationship</b>",
      "<br><br>Test performance: %{x:.1f}%",
      "<br>Predicted HAVO+ advice: %{y:.1f}%",
      "<extra></extra>"
    ),
    showlegend = FALSE
  )

# Lower- and higher-disadvantage predictions
for (group in c("Lower disadvantage", "Higher disadvantage")) {
  line_data <- prediction_interactive |>
    filter(disadvantage == group)

  interactive_plot <- interactive_plot |>
    add_lines(
      data = line_data,
      x = ~pct_test,
      y = ~fit,
      name = group,
      text = ~hover_text,
      hovertemplate = "%{text}<extra></extra>",
      line = list(color = group_colors[[group]], width = 5),
      visible = FALSE,
      showlegend = TRUE,
      legendgroup = group
    )
}

interactive_plot <- interactive_plot |>
  layout(
<<<<<<< HEAD
    margin = list(l = 75, r = 25, t = 125, b = 95),
    
    # place the title and subtitle at a fixed pixel distance above the plot
    # ensures spacing stays the same when the plot is resized 
=======
    margin = list(l = 85, r = 25, t = 225, b = 240),

>>>>>>> d0ca5faa28835aa739bfcd8bda77ac3cdbefaab4
    annotations = list(
      # Title
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
<<<<<<< HEAD
        xanchor = "left", 
        yanchor = "bottom",
        align = "left", 
        yshift = 88,
        text = "<b>The Hidden Gap between Test Performance and School Advice</b>",
        showarrow = FALSE,
        font = list(size = group_px$title)
      ),
      # --- subtitle --- 
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        yanchor = "bottom",
        yshift = 66,
        text = paste0(
          "Does school disadvantage reveal ",
          "differences that the overall trend hides?"
        ),
        showarrow = FALSE,
=======
>>>>>>> d0ca5faa28835aa739bfcd8bda77ac3cdbefaab4
        xanchor = "left",
        yanchor = "bottom",
        align = "left",
        yshift = 112,
        text = title_text,
        showarrow = FALSE,
        font = list(size = group_px$title, color = group_prim_text)
      ),

      # Dynamic subtitle (annotation index 1)
      list(
        x = 0,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "bottom",
        align = "left",
        yshift = 58,
        text = subtitle_overall,
        showarrow = FALSE,
        font = list(size = group_px$subtitle, color = group_sec_text)
      ),

      # Hover instruction
      list(
        x = 1,
        y = 1,
        xref = "paper",
        yref = "paper",
        xanchor = "right",
        yanchor = "bottom",
        yshift = 8,
        text = "Hover over the lines to compare predictions",
        showarrow = FALSE,
        font = list(size = 12, color = group_sec_text)
      ),

      # Additional information
      list(
        x = 0,
        y = 0,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        yshift = -105,
        align = "left",
        text = footnote_1,
        showarrow = FALSE,
        font = list(size = 11, color = group_sec_text)
      ),

      # Source and authors
      list(
        x = 0,
        y = 0,
        xref = "paper",
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        yshift = -180,
        align = "left",
        text = footnote_2,
        showarrow = FALSE,
        font = list(size = 11, color = group_sec_text)
      )
    ),

    xaxis = list(
      title = "Pupils reaching the target test level (%)",
      range = c(0, 100),
      tickvals = seq(0, 100, 20),
      showline = TRUE,
      automargin = TRUE
    ),

    yaxis = list(
      title = "Pupils receiving HAVO or higher advice (%)",
      range = c(0, 80),
      tickvals = seq(0, 80, 20),
      showline = TRUE,
      automargin = TRUE
    ),

    legend = list(
      orientation = "h",
      x = 0.5,
      y = 0.14,
      xanchor = "center",
      yanchor = "middle"
    ),

    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0,
        y = 1,
        xanchor = "left",
        yanchor = "bottom",
        showactive = TRUE,
        buttonwidth = 190,
        bgcolor = "#EAE1D4",
        bordercolor = "#B9AA94",
        borderwidth = 1,
        font = list(size = 14, color = group_prim_text),
        pad = list(b = 12),

        buttons = list(
          list(
            label = "Overall relationship",
            method = "update",
            args = list(
              list(
                visible = c(
                  TRUE,
                  FALSE, FALSE,
                  TRUE,
                  FALSE, FALSE
                )
              ),
              list(
                showlegend = FALSE,
                "annotations[1].text" = subtitle_overall
              )
            )
          ),

          list(
            label = "By disadvantage",
            method = "update",
            args = list(
              list(
                visible = c(
                  FALSE,
                  TRUE, TRUE,
                  FALSE,
                  TRUE, TRUE
                )
              ),
              list(
                showlegend = TRUE,
                "annotations[1].text" = subtitle_gap
              )
            )
          )
        )
      )
    )
  )

interactive_plot |> theme_interactive()

```

### Current conclusion

Schools with similar test performance do not have the same predicted HAVO-or-higher advice across levels of school disadvantage. The difference is relatively small at lower test-performance levels and becomes larger as test performance rises. The interactive visualization is the current working version of the final plot; further interpretation, context, and narrative framing can be added later without changing the analysis above.